Two Glu/Asp Residues Cooperatively Mediate an Early Step of ATP Hydrolysis in GHKL ATPases MutL and GyrB

Version 1

Abstract

GHKL ATPases share a unique Bergerat ATP-binding fold and regulate diverse biological processes through ATP-dependent conformational changes. An early step of ATP hydrolysis in this family has been attributed to a single highly conserved glutamate residue proposed to function as the general base. However, mutations of this residue impair both the ATPase activity and ATP binding, complicating interpretation of its catalytic role. Re-examination of the high-resolution crystal structures revealed a second conserved acidic residue positioned within a hydrogen-bonding distance from the nucleophilic water molecule. Using Aquifex aeolicus MutL and GyrB as model enzymes, we combined systematic mutagenesis, ATPase and ATP-binding assays, and X-ray crystallography to dissect the roles of these residues. We show that alignment of the nucleophilic water can be maintained as long as the conserved glutamate retains hydrogen-bonding capability, whereas efficient ATP hydrolysis requires proton-accepting capacity at least one of the two acidic residues. These results indicate that the conserved glutamate primarily governs positioning of the nucleophilic water, while activation of this water for catalysis is achieved through cooperative general base function of the two acidic residues. Extending this framework to human MutL homologs, PMS2 and MLH1, we showed that clinically reported variants of uncertain significance in these DNA mismatch repair proteins substantially reduced the ATPase activity, indicating functional impairment. Together, our findings refine the catalytic mechanism of GHKL ATPases and provide a structural and functional framework for interpreting disease-associated variants in GHKL ATPases. Phylogenetic and ancestral state analysis further indicated that the second acidic residue was likely to be present in the common ancestor of major GHKL ATPase lineages but was later modified in a branch including Hsp90, suggesting evolutionary remodeling of the catalytic mechanism in the branch.

Introduction

The GHKL ATPase family comprises a wide range of proteins involved in diverse cellular processes, including DNA replication, DNA repair, protein homeostasis, chromatin regulation, and signal transduction 1. The GHKL ATPase family includes the following representative members: DNA gyrase B (GyrB), the ATPase subunit of type II topoisomerase that introduces negative supercoils into DNA 2,3; the molecular chaperone Hsp90, which regulates the folding and stability of numerous client proteins 4; histidine kinases of two-component regulatory systems, which function in environmental sensing and signal transduction 5; the DNA mismatch repair protein MutL, which maintains genome stability by correcting replication errors 6,7; and MORC family proteins, which contribute to chromatin organization and epigenetic regulation 8,9. Despite their functional diversity, these proteins share a conserved Bergerat ATP-binding motif and are thought to employ a common mechanism for ATP binding and hydrolysis 1,10–13.

For GHKL ATPases, ATPase activity is essential for protein function. In many cases, ATP binding and hydrolysis drive large-scale conformational changes that are coupled to biological activity. The ATP-bound and ADP-bound states often correspond to distinct oligomeric or domain arrangements, and cycling between these states enables GHKL ATPases to act as molecular switches or motors that coordinate complex cellular processes 14–16. A representative example of such functional coupling is provided by MutL, a key component of the DNA mismatch repair pathway 17–19. MutL consists of an N-terminal ATPase domain and a C-terminal nuclease domain, and ATP binding induces dimerization of the N-terminal domains, leading to rearrangement of the overall architecture and engagement of downstream repair factors 14,20–22. Subsequent ATP hydrolysis, stimulated by interacting proteins and DNA substrates, induces further conformational changes that activate its nuclease function and enable strand incision during mismatch repair 7,23–25. Thus, in MutL, the ATP binding and hydrolysis cycle directly controls large-scale structural transitions that are essential for its biological activity. A detailed understanding of the ATP hydrolysis mechanism is therefore crucial not only for elucidating the molecular basis of fundamental biological phenomena but also for clarifying disease mechanisms associated with dysfunction of GHKL ATPases.

In human, impaired function of MutL proteins compromises the DNA mismatch repair pathway, leading to reduced fidelity of DNA replication. Such defects are a major cause of Lynch syndrome, one of the most common hereditary cancer predisposition syndromes 26–30. Cancers associated with Lynch syndrome exhibit high responsiveness to immune checkpoint inhibitors 31,32, and genetic testing of MutL genes is recommended for diagnosis. However, clinical interpretation is often hampered by the presence of variants of uncertain significance. This limitation largely reflects the insufficient understanding of MutL function at the amino acid level. Variants of uncertain significance have also been identified in the ATPase active site of human MutL homologs, highlighting the importance of elucidating the ATP hydrolysis mechanism of MutL for establishing a molecular basis for genetic diagnosis of Lynch syndrome.

Among GHKL ATPases, GyrB, MutL, and MORC are thought to share particularly similar ATPase mechanisms 11. Within the Bergerat ATP-binding motif, a highly conserved glutamate residue, Glu29 in Aquifex aeolicus MutL (aqMutL), has been proposed to function as a general base that activates the nucleophilic water molecule for ATP hydrolysis 33,34. This proposal is based on observations that substitution of alanine or lysine for this glutamate abolishes the ATPase activity. In addition, a conserved lysine residue, Lys79, in aqMutL, has been suggested to act as a general acid, donating a proton to the reaction intermediate and thereby facilitating cleavage of the γ-phosphate 11. Other conserved basic residues have also been proposed to contribute as Lewis acids that stabilize negatively-charged reaction intermediates 11,34. Collectively, these residues have been proposed to form the core catalytic machinery of GHKL ATPases.

Previously, we determined high-resolution crystal structures of the ATPase domain of aqMutL in complex with ATP analogs, in which a water molecule presumed to be the nucleophile for ATP hydrolysis was clearly observed at a hydrogen-bonding distance to the conserved glutamate Glu29 35. Upon careful re-examination of these structures, we noticed that this same water molecule is also positioned within a hydrogen-bonding distance to a second conserved acidic residue, Glu32 (Fig. 1A). Thus, the nucleophilic water is not exclusively coordinated by Glu29, but simultaneously interacts with two acidic residues.

Structural position and conservation of two acidic residues in the GHKL ATPase family.

(A) Crystal structure of the aqMutL N-terminal domain (NTD) in complex with the non-hydrolyzable ATP analog AMPPCP. The boxed region indicates the ATP-binding site and is shown enlarged on the right. AMPPCP and the side chains of Glu29 and Glu32 are shown as stick models. Oxygen atoms are shown in red, nitrogen atoms in blue, and phosphorus atoms in orange. The magnesium ion and the water molecule putatively acting as the nucleophile are depicted as green and red spheres, respectively. Hydrogen bonds are indicated by dashed lines, with distances shown. The blue mesh represents an omit Fo–Fc electron density map contoured at 3σ for AMPPNP, the magnesium ion, and the water molecule. (B) Sequence alignment of representative GHKL ATPases. The Bergerat ATP-binding fold is underlined, and the conserved acidic residues corresponding to aqMutL Glu29 and Glu32 are marked with asterisks. Amino acid sequence alignment was performed by CLUSTAL W program 57 and visualized by ESPript3 58.

Although earlier biochemical studies, including those from our group, demonstrated that replacement of Glu29 with alanine abolishes the ATPase activity, such mutations also severely impair ATP binding 11,33, complicating interpretation of the specific catalytic role of this residue. In this context, the potential contribution of Glu32 to ATP hydrolysis might have been overlooked. Supporting this notion, the residue corresponding to Glu32 is conserved not only in MutL but also in other GHKL ATPases, including GyrB and MORC proteins (Fig. 1B). Furthermore, replacements of the corresponding Glu residue with Lys in Escherichia coli MutL and in human MutL homolog 1 (MLH1) have been reported to abolish DNA mismatch repair activity in cell-based or nuclear extract-based assays 36,37.

In this study, we investigated the functional role of Glu32 in the ATP hydrolysis mechanism of MutL and GyrB using a combination of enzymatic and X-ray crystallographic analyses. Our results provide new insights into the catalytic mechanism of GHKL ATPases and refine the current model of ATP hydrolysis for this protein family. By clarifying the molecular basis of ATP hydrolysis in GHKL ATPases, this work also provides a mechanistic framework for interpreting the functional consequences of some clinically observed variants, thereby contributing to the foundation for assessing the impact of variants of uncertain significance.

Results and Discussion

Mutational analysis of Glu29 and Glu32 in the ATPase domain of aqMutL

To examine the contribution of Glu32 to the ATPase activity of the aqMutL NTD and to directly compare its role with that of Glu29, we introduced a series of substitutions at both positions. To exclude the possibility that the observed effects on the ATPase activity arose from global structural perturbations, circular dichroism (CD) spectra were measured for all mutant forms (Fig. 2A). Among these, only the E32G mutant form exhibited a marked decrease in ellipticity, suggesting that this substitution disrupts the secondary structure of the protein. In contrast, all other mutant forms displayed CD spectra indistinguishable from that of the wildtype form, indicating that these substitutions do not substantially perturb the overall structure of the aqMutL NTD.

Effects of mutations on the overall structure and the ATPase activity of the aqMutL NTD.

(A) Far-UV CD spectra of the wildtype and mutant forms of the aqMutL NTD were collected and averaged over 10 consecutive scans. (B, C) ATPase activity of the aqMutL NTD was measured under steady-state conditions using a colorimetric assay that detects the release of inorganic phosphate. Apparent rate constants (kapp) were calculated from the measured phosphate concentrations and plotted as a function of ATP concentration. Kinetic parameters were determined by fitting Michaelis–Menten equation to the data. Each data point represents the mean of three independent experiments, with error bars indicating standard deviation. Theoretical curves calculated from the fitted kinetic parameters are overlaid. Data for 0–1 mM substrate concentrations are shown to compare the wildtype from with mutant forms (B). Activities for higher substrate concentrations were measured for some of the mutant forms (C).

We first analyzed the effects of mutations at Glu29 on the ATPase activity of the aqMutL NTD. The mutation to Gln would selectively impair the proton association/dissociation capability of the Glu side chain while preserving its size and hydrogen-bonding potential. Moreover, because mutations to Val, Lys, and Gly at this position have been reported as variants of uncertain significance in human MutL, these mutations were also analyzed. Consistent with previous reports, all Glu29 mutant forms exhibited severe reduction in the ATPase activity (Fig. 2B and C, and Table 1). In particular, the E29A, E29V, E29K, and E29G mutant forms showed activities below the detection limit, indicating that Glu29 plays a central role in ATP hydrolysis. However, as described above, previous studies also demonstrated that mutations at Glu29 impair ATP binding 11,33,34, indicating that the loss of the ATPase activity of these mutant forms cannot be unambiguously attributed to disruption of a catalytic base function.

To further dissect the role of Glu29, we focused on the E29Q mutant form. Using equilibrium dialysis, we confirmed that the E29Q mutant form binds ATP with an affinity comparable to that of the wildtype protein (Supplementary Fig. S1A). Because this substitution does not impair ATP binding, it should enable a more direct assessment of the proposed role of Glu29 as a general base in ATP hydrolysis. The E29Q mutant form retained measurable ATPase activity, although its kcat value was approximately tenfold smaller than that of the wildtype enzyme (Fig. 2B and C, and Table 1). The persistence of substantial catalytic activity in the absence of the carboxylate group at the position 29 suggests that a residue other than Glu29 can participate in proton abstraction from the nucleophilic water during ATP hydrolysis.

We next examined the effects of mutations at Glu32. Mutations of Glu32 resulted in reduced ATPase activity relative to the wildtype form, supporting a role for this residue in catalysis. However, with the exception of the E32K and the structurally perturbed E32G mutant forms, all Glu32 mutant forms retained detectable ATPase activity (Fig. 2B and C, and Table 1), indicating that loss of Glu32 alone is insufficient to abolish ATP hydrolysis. Simultaneous substitutions of glutamine for the two residues in the E29Q/E32Q double mutant form led to a complete loss of the ATPase activity. Taken together, these results indicate that Glu29 and Glu32 do not exert equivalent functions but act cooperatively. The presence of at least one carboxylate group at either position, 29 or 32, appears to be required for ATP hydrolysis.

Conserved roles of two acidic residues of aqGyrB in ATP hydrolysis

To examine whether the cooperative roles of Glu29 and Glu32 observed in aqMutL are conserved among other GHKL ATPases, we next analyzed the NTD of A. aeolicus GyrB (aqGyrB). A glutamate residue corresponding to aqMutL Glu29 is also conserved in bacterial GyrB and has been proposed to function as a general base in ATP hydrolysis 38. Although crystal structure of the E. coli GyrB NTD has been reported 39, the position of the putative nucleophilic water has not been clearly defined. Therefore, we selected the GyrB NTD from the hyperthermophilic bacterium A. aeolicus as a model system, anticipating that it would allow high-resolution structural analysis and clearer visualization of a catalytic water molecule.

First, we determined the crystal structure of this protein in complex with AMPPNP at 1.65 Å resolution (Table 2 and Fig. 3). The overall structure was highly similar to that of the E. coli GyrB NTD, with a Cα root mean square deviation of 1.42 Å. As observed for the E. coli GyrB NTD, AMPPNP was accommodated by the Bergerat ATP-binding fold, and AMPPNP binding seemed to induce conformational changes that promoted dimer formation through the newly formed interface. Glu48 and Asp51 of aqGyrB correspond to Glu29 and Glu32 of aqMutL, respectively. In the crystal structure, these residues were found to tightly coordinate a water molecule that is likely to act as the nucleophilic water during ATP hydrolysis (Fig. 3). Based on these structural observations, we introduced Ala or Gln/Asn substitutions for Glu48 and Asp51. The CD spectra of the mutant forms were nearly identical to that of the wildtype form (Fig. 4A), suggesting that these substitutions did not disrupt the overall structure of the aqGyrB NTD.

Structure of the AMPPNP-bound dimer of the wildtype aqGyrB NTD.

The boxed region highlights the ATP-binding site and is enlarged in the left panel. The same region is further enlarged and rotated 60° to provide an alternative view of the active site geometry. AMPPNP and the side chains of Glu48 and Asp51 are shown as stick models. Oxygen atoms are shown in red, nitrogen atoms in blue, and phosphorus atoms in orange. The magnesium ion and the water molecule presumed to act as the nucleophile are depicted as green and red spheres, respectively. Hydrogen bonds are indicated by dashed lines, with distances shown. The blue mesh represents an omit Fo–Fc electron density map contoured at 3σ for AMPPNP, the magnesium ion, and the water molecule.

Effects of mutations on the overall structure and the ATPase activity of the aqGyrB NTD.

(A) Far-UV CD spectra of the wildtype and mutant forms of the aqGyrB NTD were recorded and averaged over 10 accumulations. (B) ATPase activity of the aqGyrB NTD was measured under steady-state conditions as described in Fig. 2. The kapp values were plotted as a function of ATP concentration. Michaelis–Menten equation was fitted to the data. Data points represent the mean ± standard deviations from three independent experiments.

Equilibrium dialysis showed that the E48A mutant form failed to bind ATP, whereas the E48Q mutant form displayed ATP-binding affinity comparable to that of the wildtype protein (Supplementary Fig. S1B). The E48Q mutation therefore allowed us to evaluate the catalytic contribution of Glu48 independently of ATP binding. ATPase activity assays showed that the E48A mutant form completely lacked the activity, while the E48Q mutant form retained approximately one quarter of the wildtype activity (Fig. 4B and Table 1). The persistence of substantial catalytic activity in the E48Q mutant form suggests that another residue partially compensates for the loss of the Glu48 carboxylate in the proton abstraction. Mutations of Asp51 also impaired ATP hydrolysis, with the D51A and D51N mutant forms retaining ~25% and ~50% of the wildtype activity, respectively (Fig. 4B and Table 1). Notably, the E48Q/D51N double mutant form exhibited no detectable ATPase activity, resembling the effect of the corresponding double mutation in the aqMutL NTD. Taken together, these results indicate that both Glu48 and Asp51 contribute to the catalytic mechanism of ATP hydrolysis in aqGyrB and that at least one of these residues must retain a carboxylate group with the proton-accepting capability to sustain the enzymatic activity.

Crystal structures of aqGyrB mutant forms and the role of Glu48 in ATP binding

In order to relate the ATPase activity profiles of the mutant forms to their structural features, we determined X-ray crystal structures of selected mutant forms using the aqGyrB NTD as a model system, which allows relatively high-resolution structural analyses. The AMPPNP-bound structures of the E48A, E48Q, D51A, and D51N mutant forms were determined at resolutions of 2.66, 1.70, 1.63, and 2.05 Å, respectively (Table 2).

Since the conserved general base Glu (i.e. Glu48 of aqGyrB) does not directly interact with ATP, it has been unclear why mutations at the Glu residue abolish the ATP-binding ability of GHKL ATPases. To clarify this, we compared the crystal structures of the wildtype and mutant forms of the aqGyrB NTD (Fig. 5). In GHKL ATPases, ATP binding is tightly coupled to conformational rearrangements, including formation of an ATP lid (Fig. 5A), which acts to secure the bound nucleotide by enclosing it within the active site 3,21. Structural analysis of the wildtype aqGyrB NTD revealed that the side chain of Glu48 forms a hydrogen bond with Gln340, thereby stabilizing a loop containing Gln340 in a conformation capable of ATP binding (Fig. 5B). In this arrangement, Gln340 in turn forms a hydrogen bond with His122, a residue located within the ATP lid, while Lys342 in the same loop is positioned to interact with the γ-phosphate of ATP. These coordinated interactions appear to promote the ATP-induced conformational transition required for stable nucleotide binding.

Structural basis for the loss and preservation of ATP binding in aqGyrB NTD mutant forms.

The wildtype and all mutant forms were crystallized in the presence of AMPPNP; however, the E48A mutant form did not bind AMPPNP, whereas the E48Q, D51A, and D51N mutant forms were observed in complex with AMPPNP. (A) Overall structures of the wildtype and E48A mutant aqGyrB NTD monomers. The ATPase active site is indicated by a boxed region. The ATP-lid (residues 105–124) in the wildtype form and the corresponding region in the E48A mutant form are shown in magenta and olive, respectively. (B–F) Enlarged views of the ATPase active site in the wildtype (B), E48A (C), E48Q (D), D51A (E), and D51N (F) forms. Side chains of the residues implicated in the ATP-induced conformational changes (e.g. His122, Gln340, and Lys342), as well as residues 48 and 51, are shown as stick models. AMPPNP is depicted as sticks, and the Mg2+ ion is shown as a green sphere. Hydrogen bonds and ionic interactions are indicated by dashed lines.

In contrast, structural analysis of the E48A mutant form demonstrated that disruption of the interaction with Gln340 abolishes this network of hydrogen bonds, preventing the associated conformational rearrangements and thereby impairing ATP binding (Fig. 5C). These findings indicate that the interaction network emanating from Glu48 is indispensable for the conformational changes associated with the ATP lid formation. Interestingly, in the E48Q mutant form, the hydrogen bond with Gln340 is preserved, allowing the conformational transition to occur and enabling ATP binding (Fig. 5D). Consistently, in the D51A and D51N mutant forms, the interaction mediated by Glu48 remains intact, and the structural rearrangements required for ATP binding are maintained (Fig. 5E and F).

These structural observations extend previous models of GyrB ATPase function by linking the conserved catalytic glutamate to ATP lid stabilization and nucleotide-dependent conformational rearrangement. The E48A aqGyrB NTD structure, in particular, provides direct structural evidence that disruption of this Glu-centered interaction network is sufficient to prevent ATP-induced active-site assembly. This hydrogen-bonding network emanating from the general base Glu is conserved in GyrBs from other organisms 39,40 (Supplementary Fig. S2), while it is not observed in other GHKL ATPases, including MutL 22,34,35,41 and MORC family proteins 9,42,43. Therefore, the loss of ATP-binding ability observed upon the mutation of Glu29 to Ala in MutL 11 is likely to be explained by a different mechanism. This difference highlights mechanistic diversity within the GHKL ATPase family, despite conservation of the two catalytic acidic residues.

Nucleophilic water molecule in the active site of mutant forms of the aqGyrB NTD

Having established the structural basis for the altered ATP-binding properties of the mutant forms, we next analyzed how these substitutions influence the positioning of the nucleophilic water molecule. Structural comparisons of the catalytic sites of the four mutant forms are shown in Fig. 6. In the E48A mutant form (Fig. 6A), the nucleophilic water molecule was not observed in the catalytic site. In contrast, in the E48Q mutant form, the nucleophilic water molecule was positioned at essentially the same location as in the wildtype form, coordinated by Gln48 and Asp51, and bound AMPPNP was observed (Fig. 6B). This structural arrangement is consistent with the biochemical data described above, in which the E48Q mutant form retained substantial ATP-binding and ATPase activities.

Structures surrounding the conserved acidic residues: (A) E48A, (B) E48Q, (C) D51A, and (D) D51N mutant forms of the aqGyrB NTD.

AMPPNP and the side chains at the positions 48 and 51 are shown as stick models. Oxygen atoms are shown in red, nitrogen atoms in blue, and phosphorus atoms in orange. The magnesium ion is shown as a green sphere, and the water molecule proposed to act as the nucleophilic water is shown as a red sphere. Blue mesh represents Fo–Fc omit electron density maps contoured at 3σ, calculated after omission of AMPPNP, the magnesium ion, and the nucleophilic water molecule. Hydrogen bonds are indicated by dashed lines, with the corresponding distances shown.

Unlike the E48A mutant form, the D51A mutant form bound AMPPNP (Fig. 6C), and the nucleophilic water molecule was coordinated only by Glu48. This asymmetric coordination suggests that Glu48 plays a dominant role in positioning the nucleophilic water molecule. In the D51N mutant form (Fig. 6D), which lacks the proton-transfer capability of Asp51 while retaining its hydrogen-bonding capacity, the nucleophilic water was coordinated by both Glu48 and Asn51. These structural features are consistent with the relative ATPase activities observed for the mutant forms of these two acidic residues (Fig. 4B).

Taken together, these results indicate that Glu48 and Asp51 make distinct yet cooperative contributions to ATP hydrolysis. Glu48 appears to play a dominant role in aligning the nucleophilic water molecule and activating this water molecule through proton abstraction as a general base. In contrast, Asp51 seems to contribute less to the precise positioning of the nucleophilic water but participates cooperatively with Glu48 in the proton abstraction, thereby supporting efficient ATP hydrolysis. Notably, an acidic residue corresponding to Asp51 of aqGyrB is conserved among several GHKL ATPases, including MutL, GyrB, and MORC family proteins (Fig. 1B), suggesting that the two-acidic-residue mechanism proposed here may be shared among these subgroups. The second acidic residue is not conserved in other members of the GHKL family, such as Hsp90 and mitochondrial histidine kinases (Fig. 1B), where a single acidic residue is thought to fully function as the general base responsible for water activation. A similar divergence appears to apply to the previously-reported general acid Lys residue 11, further supporting the notion that GHKL ATPases can be divided into at least two mechanistically distinct subclasses. This mechanistic bifurcation provides insights into the molecular evolution of the GHKL ATPase family.

Functional and clinical implications of the second acidic residue in human MutL homologs

Our analyses of the aqMutL NTD and aqGyrB NTD indicate that, in addition to the Glu residue traditionally considered to function as the general base, an additional conserved acidic residue (Glu or Asp) also participates in the general base catalysis. In human MutL homologs, this residue corresponds to Glu44 in Postmeiotic Segregation Increased 2 (PMS2) and Glu37 in MLH1. Notably, variants of uncertain clinical significance, including E44V (ClinVar variation ID: 232732), E44G (230139), and E44Q (186069) in PMS2, as well as E37G (230139) and E37K (89640) in MLH1, have been reported at these positions.

To examine whether these residues similarly contribute to the ATPase activity in human MutL homologs, we introduced the corresponding mutations into the NTDs of human PMS2 and MLH1 and measured their ATPase activities. The recombinant E44G human PMS2 NTD and E37G human MLH1 NTD could not be expressed successfully, which is consistent with our finding that the corresponding mutation in the aqMutL NTD disrupted the structural integrity of the protein (Fig. 2A). Both the E44V and E44Q mutant forms of the PMS2 NTD exhibited a marked reduction in the ATPase activity, retaining approximately one-sixth of the wildtype activity (Fig. 7A and Table 1). In contrast, the E37K mutation in the MLH1 NTD completely abolished the ATPase activity under our assay conditions (Fig. 7B and Table 1).

ATPase activities of human MutL homolog NTDs.

(A) ATPase activity of mutant forms of the human PMS2 NTD. (B) ATPase activity of a mutant form of the human MLH1 NTD. ATPase activities were measured using the same procedures as described for Fig. 2 and Fig. 4. Apparent rate constants (kapp) were plotted as a function of substrate concentration. Data points represent the mean values from three independent measurements, with error bars indicating standard deviations. Solid lines represent the theoretical Michaelis-Menten curves.

Maximum-likelihood phylogeny of the ATPase domain of representative GHKL family members.

The state of the amino acid residue at the position corresponding to aqMutL Glu32 or aqGyrB Asp51 is indicated by colored bars (red, acidic residue; blue, non-acidic residue). The phylogenetic tree was visualized and annotated using the Interactive Tree of Life (iTOL) web server 56. The tree is displayed in a rectangular layout for clarity and should be interpreted as unrooted. Ancestral state reconstruction was performed on the fixed topology, and posterior probabilities are shown for selected internal nodes. The internal node uniting the MutL/GyrB/MORC/Hsp90 clades strongly supports an acidic residue with the posterior probability (PP) of 0.983, whereas the Hsp90 stem ancestor is strongly supported to encode serine at this position with the posterior probability of 1.000. Species abbreviations used in the phylogenetic tree are as follows: Ce, Caenorhabditis elegans; Sc, Saccharomyces cerevisiae; Dr, Danio rerio; Hs, Homo sapiens; Aa, Aquifex aeolicus; Hp, Helicobacter pylori; Tt, Thermus thermophilus; Ec, Escherichia coli; Mt, Mycobacterium tuberculosis; Bs, Bacillus subtilis; Sm, Streptococcus mutans; Dm, Drosophila melanogaster; Clf, Canis lupus familiaris; Mm, Mus musculus; Hc, Hemicordylus capensis; Pv, Patella vulgata; Tf, Tachysurus fulvidraco; Lk, Lepidochelys kempii; Cc, Castor canadensis; Dn, Dasypus novemcinctus.

These results are in good agreement with our observations for the aqMutL NTD and aqGyrB NTD, further supporting the notion that this second acidic residue plays a conserved and functionally significant role in ATP hydrolysis across the GHKL ATPase family. Previous studies demonstrated that even a several-fold reduction in the MutL ATPase activity could severely compromise mismatch repair efficiency in vivo 11. In this context, the substantial reduction in the ATPase activity caused by the E44V and E44Q mutations in the PMS2 NTD is likely to impair mismatch repair in human cells. Therefore, these variants might contribute to the development of Lynch syndrome by weakening the ATPase-driven regulatory functions of MutL. Nevertheless, whether these variants indeed compromise mismatch repair activity in vivo remains to be directly tested. Functional validation using cell-based or nuclear extract-based mismatch repair assays, such as the 6-thioguanine sensitivity assay 44 or CIMRA assay 36, will be required to establish a definitive link between the reduced ATPase activity and mismatch repair deficiency in human cells.

Phylogenetic analysis of GHKL ATPases

The newly-identified functionally-important acidic residue is not uniformly conserved across the GHKL superfamily. We therefore used this position as an evolutionary marker to investigate patterns of molecular divergence within the GHKL ATPase domain. In histidine kinases, even the canonical catalytic glutamate is typically replaced, and the ATPase-associated catalytic machinery is not maintained 1,45,46. Therefore, phylogenetic reconstruction was restricted to other GHKL ATPase family members MutL, GyrB, MORC, and Hsp90.

Phylogenetic analysis based on the trimmed ATPase-domain alignment recovered these major GHKL ATPase lineages (MutL, GyrB, MORC, and Hsp90) as a well-supported cluster, indicating relatively close evolutionary relationships among these groups, which is consistent with a previous study 1. Ancestral state reconstruction performed on the maximum-likelihood phylogeny strongly supports the presence of an acidic residue at the position corresponding to aqMutL Glu32 or aqGyrB Asp51 in the internal node uniting the MutL, GyrB, MORC, and Hsp90 clades with a posterior probability of 0.983. In contrast, the ancestral node of the Hsp90 clade is strongly supported to encode serine at this position with a posterior probability of 1.000. Because no external outgroup was specified, the global polarity of amino acid residue changes across the entire phylogeny cannot be formally determined. Nevertheless, the widespread conservation of the second acidic residue across the MutL, GyrB, and MORC lineages, together with the high posterior probability at the most recent common ancestor of these groups and Hsp90, indicates that this residue was very likely to be present in their ancestor. Its absence within the Hsp90 lineage is therefore most parsimoniously interpreted as lineage-specific loss. In summary, these results suggest that the Hsp90 lineage underwent remodeling of its ATPase catalytic mechanism during its evolutionary divergence from other GHKL ATPase family members.

Conclusions

In this study, we re-examined the catalytic mechanism of GHKL ATPases using aqMutL and aqGyrB as model systems with mutational, biochemical, and crystallographic studies. Our results reveal that ATP hydrolysis in these enzymes is not governed by a single general base residue, as has been traditionally assumed, but instead relies on the coordinated action of two conserved acidic residues. Structural analyses indicate that Glu48 of aqGyrB plays a dominant role in aligning the nucleophilic water molecule in an optimal geometry for attack on the γ-phosphate of ATP. However, biochemical and structural evidence from single and double mutant forms indicates that both residues contribute to proton abstraction from the nucleophilic water. Thus, general base catalysis in GyrB and MutL appears to be achieved by redundant proton-accepting residues.

Extending this finding to human MutL homologs, we showed that mutations at the corresponding residues in PMS2 and MLH1—reported as variants of uncertain clinical significance—lead to substantial or complete loss of the ATPase activity. Given that even modest reductions in the MutL ATPase activity are known to severely compromise mismatch repair efficiency in vivo, our findings suggest that these variants might be functionally deleterious. Thus, the mechanistic insights obtained in this study provide a structural and biochemical basis for evaluating the pathogenic potential of clinically ambiguous variants in human MutL homologs.

Materials and Methods

Construction of expression plasmids

Expression plasmids encoding mutant forms of the aqMutL NTD (residues 1–315), the aqGyrB NTD (residues 1–397), the human PMS2 NTD (residues 1–365), and the human MLH1 NTD (residues 1– 337) were generated by introducing point mutations into pET-11a/aqMutL NTD 47, pET-11a/aqGyrB NTD 11, pCold ProS2/human PMS2 NTD 35, and pET28-MLH1/human MLH1 NTD 11 plasmids using the PrimeSTAR site-directed mutagenesis protocol (Takara). Primer sequences used for mutagenesis are listed in Supplementary Table S1. DNA sequencing confirmed that no unintended mutations were introduced.

Expression and purification of proteins

The wildtype aqMutL NTD, aqGyrB NTD, ProS2-tagged human PMS2 NTD, and His-tagged human MLH1 NTD were overexpressed and purified as described previously 11,35,47. The mutant forms were overexpressed and purified using the same procedures as for the respective wildtype proteins.

CD spectrometry

CD measurements were performed with a spectropolarimeter, model J-720W (Jasco). Measurements of CD spectra were carried out in a solution comprised of 20 mM Tris-HCl (pH 8.0) (eLANT) and 10 μM protein using a 0.1 cm cell at 25°C. The residue molar ellipticity [Θ] was defined as 100 Θobs/(lc), where Θobs is the observed ellipticity, l is the length of the light path in centimeters, and c is the residue molar concentration of the protein.

ATPase assay

ATPase activities of the aqMutL NTD, aqGyrB NTD, ProS2-tagged human PMS2 NTD, and His-tagged human MLH1 NTD were measured as described previously 35. Reactions were carried out using 10 μM protein in a 20 μl reaction mixture containing 50 mM HEPES–KOH (pH 7.5) (eLANT), 100 mM NaCl, 5 mM MgCl2, and various concentrations of ATP (Roche). For the aqMutL NTD and aqGyrB NTD, reactions were performed at 70°C for 60 min. For the NTDs of the human MutL homologs, reactions were performed at 37°C for 120 min. All experiments were conducted in triplicate. Kinetic parameters were determined by fitting the data to the standard Michaelis–Menten equation using IgorPro.

Equilibrium dialysis

Equilibrium dialysis experiments were performed as described previously 11. Seventy-five microliters of buffer containing 50 mM Tris–HCl (pH 8.0), 100 mM NaCl, 5 mM MgCl2, and 50 μM AMPPNP (Roche) were loaded into the buffer chamber of a DispoEquilibrium Dialyzer equipped with a 10 kDa molecular weight cut-off membrane (Harvard Bioscience, Inc.). An equal volume of the aqMutL NTD (0–320 μM) or aqGyrB NTD (0–400 μM) in the same buffer without AMPPNP was placed in the opposing sample chamber. The device was gently agitated at room temperature for 24 h to allow equilibration. The absorbance of the solution in the buffer chamber was measured at 260 nm. The concentration of unbound AMPPNP was calculated using a molar extinction coefficient (ε260 = 15,400 M−1 cm−1).

Crystallization and structure determination

Protein crystallization was performed by the sitting-drop vapor diffusion method at 20°C. One microliter of the aqGyrB NTDs (14 mg/ml) containing 0.6 mM AMPPNP (Roche) and 1 mM MgCl2 was mixed with an equal volume of the reservoir solution and equilibrated against 500 μl of reservoir solution. For the wildtype and D51A mutant forms of the aqGyrB NTD, crystals were obtained using a reservoir solution containing 0.2 M ammonium acetate, 0.1 M HEPES (pH 7.5), and 45% (v/v) 2-methyl-2,4-pentanediol. The E48A mutant form was crystallized in 0.05 M cadmium sulfate hydrate, 0.1 M HEPES (pH 7.5), and 20% (w/v) polyethylene glycol 3,350. The E48Q mutant form was crystallized under identical conditions consisting of 30% (w/v) polyethylene glycol 400, 100 mM HEPES–NaOH (pH 7.5), and 200 mM sodium chloride. The D51N mutant form was crystallized in 0.1 M sodium malonate (pH 7.0) and 12% (w/v) polyethylene glycol 3,350. Crystals of the wildtype and E48A mutant forms were soaked in the respective reservoir solutions supplemented with 30% (v/v) glycerol before being flash-cooled in liquid nitrogen. Crystal of the D51N mutant form was transferred to the reservoir solution containing 30% (w/v) polyethylene glycol 3,350 prior to flash-cooling in liquid nitrogen. Crystal of the D51A mutant form was directly flash-cooled in liquid nitrogen without additional cryoprotection.

X-ray diffraction data were collected at beamline BL45XU in SPring-8 (Hyogo, Japan) at a wavelength of 1.000 Å and a temperature of −173°C. Data were automatically collected and processed using the ZOO system 48, in which data processing was assisted by KAMO 49. The linearity of the Wilson plot was confirmed over the resolution range employed. Phases were determined by molecular replacement using Phaser with a search model predicted by AlphaFold2 50. Model building and manual correction were performed using COOT 51, and structural refinement was carried out with PHENIX refine 52. Data collection and refinement statistics are summarized in Table 2. All figures illustrating the protein structures were generated using PyMOL (Schrödinger).

Phylogenetic analysis and Ancestral state reconstruction

The ATPase domains of representative GHKL superfamily members (MutL, GyrB, MORC, and Hsp90) were aligned using amino acid sequences extracted based on domain boundaries defined by structural correspondence. The multiple sequence alignment was trimmed using trimAl 53 to remove ambiguously aligned regions. Maximum-likelihood phylogenetic inference was performed using IQ-TREE (v3.0.1) 54,55. Ancestral state reconstruction was performed on the fixed maximum-likelihood topology using IQ-TREE under the empirical Bayesian framework implemented with the selected amino acid substitution model. The state of the amino acid residue at the alignment position corresponding to aqMutL Glu32 or aqGyrB Asp52 was examined. Maximum-likelihood phylogenies were visualized and annotated using the Interactive Tree of Life (iTOL) web server 56. Trees were displayed in a rectangular layout without specification of an external outgroup. For visualization purposes, branch lengths were not scaled and were displayed with equal lengths.

eLife Assessment

In this solid work, Fukui et al. re-examined the ATP hydrolysis mechanism in GHKL ATPases, revealing a cooperative role for two conserved acidic residues rather than a single one. This useful study used a range of biochemical and structural techniques on various mutants from different members of the GHKL ATPase family to test and validate their proposed mechanism. An updated and extended mechanistic model of ATP hydrolysis by this class of enzymes is proposed.

Reviewer #1 (Public review):

Summary:

In this manuscript, the authors study two residues in the GHKL ATPase active site of Aq MutL and GyrB, and argue that the catalytic base function is shared between two conserved acidic residues that are 3 residues apart.

They generated mutant versions in MutL and GyrB (both ala and the appropriate Asn/Gln version) and performed ATPase analysis. They also generated high-resolution crystal structures of the GyrB NTD with AMPPnP for WT and mutants of the two acidic residues. The data show that mutation in either of these residues does not fully kill activity (with the exception of the Alanine mutation of the first of the two, which interferes with ATP (or AMPPnP) binding). When the acidic residues are mutated to Asn/Gln, the catalytic water can still be positioned, and hence these mutants are more active than the Ala mutants. In both cases, the double mutation is catalytically dead.

The authors then perform phylogenetic analysis and ancestral gene reconstruction, and based on this, they argue that HSP90 forms a different class of GHKL ATPases, and lost rather than gained this separate status.

Strengths:

The biochemical analysis seems solid.

Weaknesses:

(1) A major question that remains is why the mutations have so much more detrimental effect in MutL (100-fold lower kcat/KM) than they do in GyrB (3-fold lower). Can the authors explain this? Doesn't this argue against the proposed catalytic conservation?

(2) The structure figures all have omit maps for just the AMPPnP and the water, whereas the density for the acidic residues and their mutants is not shown.

Reviewer #2 (Public review):

Summary:

In this manuscript, Fukui et al. re-examined the ATP hydrolysis mechanism in GHKL ATPases, revealing a cooperative role of two conserved acidic residues rather than one. The authors have used a range of biochemical and structural techniques on various mutants from different members of the GHKL ATPase family to test and validate their proposed mechanism.

Through a detailed re-analysis of their previously published structure of the aqMutL NTD (ATPase domain) in complex with AMPPCP, they identified Glu29 and Glu32 as interacting with nucleophilic water for the catalysis. The authors carefully dissected the respective roles of these two acidic residues with a series of site-directed mutations. Mutations at Glu29 impaired ATPase activity without affecting protein secondary structure or ATP binding in the case of the E29Q mutant. Moreover, mutations at Glu32 did not affect secondary structure (except for E32G) but reduced ATPase activity. Activity was abolished when both residues (E29Q/E32Q) were mutated.

The authors extended their study to another GHKL ATPase, aqGyrB. Their findings further supported the cooperative function of the corresponding acidic residues in aqGyrB (Glu48 and Asp51) during ATP hydrolysis. Mutation of these residues partially impaired ATP hydrolysis without affecting protein secondary structure. ATPase activity was completely lost in the double mutant E48Q/D51M. While the E48Q mutant retained the ability to bind ATP, the E48A mutant did not. High-resolution structures of the WT and E48A, E48Q, D51A, and D51N mutants of the aqGyrB NTD demonstrated that nucleophilic water positioning depended on these residues. E48 played a dominant role in water positioning and is critical for stabilising ATP lid formation and associated conformational changes, whereas D51 contributed cooperatively to catalysis.

The authors investigated the functional impact of mutating the corresponding residues in the human MutL homologs PMS2 and MLH1. Clinical variants consistently exhibited reduced or abolished ATPase activity, providing a potential molecular basis for Lynch syndrome through impaired DNA mismatch repair.

Lastly, through evolutionary analysis, the authors inferred that the second acidic residue was likely present in the common ancestor of MutL, GyrB, and MORC proteins, but was lost in the case of Hsp90.

Strengths:

(1) This study contains a detailed structural and biochemical analysis of a biologically important set of GHKL ATPases. The authors identify a second acidic residue that is conserved and contributes to catalysis in a large subset of GHKL ATPases. An updated and extended mechanistic model of ATP hydrolysis by this class of enzymes is proposed, which involves cooperative and partially overlapping roles for the catalytic residue pair. This revised mechanistic model is invaluable for the interpretation of clinical variants of GHKL ATPases such as PMS2 and MLH1.

(2) The work described was performed to an excellent and rigorous technical standard. The structural and biochemical data are sound. The evidence supporting the claims is compelling.

Weaknesses:

(1) The identification in this study of a second acidic residue contributing to catalysis but not absolutely essential for catalysis is a useful finding. However, given that many structures of GHLK ATPases have been determined with different nucleotide analogs bound and that the essential role of the first acidic residue is well established, the importance and scope of the advances described here remain focused within the field of study of GHKL ATPases.

(2) The authors assessed the consequences of variants in the human MutL homologs PMS2 and MLH1, but various other human GHKL ATPases contain clinically relevant variants, some of which have stronger disease associations than the mutations examined in this study. A broader analysis of the effect (or likely effect) of disease-linked mutations in GHKL ATPases would have strengthened this study.

(3) In MLH1, the E37K mutation completely abolishes ATPase activity, but the corresponding mutations in aqMutL, aqGyrB, and PMS2 do not. It remains unclear why E37K in MLH1 leads to complete loss of activity, as the authors propose that water molecule positioning via the first acidic residue, as well as ATP lid stabilisation and associated conformational changes, should still be possible.

(4) The authors do not examine ATP binding in the E32 mutants of aqMutL NTD and the D51 mutants of aqGyrB, or AMPPNP binding of the NLH1 and PMS2 mutants. Hence, the relative contributions of the acidic residues to ATP binding and hydrolysis remain partially unclear.

(5) The ATPase assays for PMS2 and MLH1 (Figure 7 and Table 1) were performed with purification/solubility tags still present. Hence, it cannot be ruled out that these tags influence the measured activities.

(6) The authors suggest that the two-acidic-residue mechanism proposed in this study could be shared among several GHKL ATPase families, yet they also state that the hydrogen-bonding network was not observed in MutL and MORC family proteins. This raises doubt about how conserved the mechanism is, e.g., in MutL and MORC proteins.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

(1) A major question that remains is why the mutations have so much more detrimental effect in MutL (100-fold lower kcat/KM) than they do in GyrB (3-fold lower). Can the authors explain this? Doesn't this argue against the proposed catalytic conservation?

We agree that the quantitative effects of the mutations differ between MutL and GyrB. However, we do not think that this difference argues against conservation of the catalytic mechanism. The trends of the mutational effects are highly consistent between the two enzymes. In both proteins, replacement of the conserved catalytic glutamate with Ala (E29A in aqMutL and E48A in aqGyrB) abolished ATPase activity and ATP binding, whereas replacement with the isosteric amide residue (E29Q and E48Q), which preserves hydrogen-bonding capability but lacks proton-accepting capacity, retained ATP binding and measurable ATPase activity. Likewise, substitutions of the second acidic residue (E32Q in aqMutL and D51N in aqGyrB) also retained substantial ATPase activity despite the loss of proton-accepting capability. Most importantly, simultaneous substitution of both acidic residues (E29Q/E32Q in aqMutL and E48Q/D51N in aqGyrB) completely abolished ATPase activity in both enzymes. Therefore, although the magnitude of the activity reduction caused by the individual mutations differs between MutL and GyrB, the qualitative pattern is essentially identical. We therefore think that the proposed catalytic mechanism is conserved, while the quantitative differences likely reflect differences in the local catalytic environment rather than differences in the underlying mechanism.

(2) The structure figures all have omit maps for just the AMPPnP and the water, whereas the density for the acidic residues and their mutants is not shown.

We have added Supplementary Fig. S2, which shows the 2Fo–Fc electron density maps around residues Glu48/Asp51 (or their substituted residues) in the wildtype and all mutant structures. The following sentences have been added in the revised manuscript:

“To show that the introduced substitutions were unambiguously supported by the crystallographic data, electron density maps around residues 48 and 51 are shown in Supplementary Fig. S2.” (p. 4 line 198-200 in the revised manuscript)

Reviewer #2 (Public review):

(1) The authors assessed the consequences of variants in the human MutL homologs PMS2 and MLH1, but various other human GHKL ATPases contain clinically relevant variants, some of which have stronger disease associations than the mutations examined in this study. A broader analysis of the effect (or likely effect) of disease-linked mutations in GHKL ATPases would have strengthened this study.

We agree that extending the analysis to additional disease-associated variants in other human GHKL ATPases would further strengthen our understanding of the conserved catalytic mechanism and its clinical relevance. However, we believe that such a comprehensive analysis is beyond the scope of the present study, which focuses on establishing the fundamental catalytic mechanism shared between MutL and GyrB. We consider systematic functional and structural analyses of disease-associated variants across the GHKL ATPase family to be an important direction for future research. We have now added a statement to the Results and Discussion section to acknowledge this limitation and highlight this future perspective:

“Although we focused here on pathogenic variants in the MutL homologs MLH1 and PMS2, extending similar structural and biochemical analyses to disease-associated variants in other human GHKL ATPases will be important for evaluating the generality and clinical relevance of the conserved catalytic mechanism proposed in this study.” (p. 6 line 303-306 in the revised manuscript)

(2) In MLH1, the E37K mutation completely abolishes ATPase activity, but the corresponding mutations in aqMutL, aqGyrB, and PMS2 do not. It remains unclear why E37K in MLH1 leads to complete loss of activity, as the authors propose that water molecule positioning via the first acidic residue, as well as ATP lid stabilisation and associated conformational changes, should still be possible.

We agree that the complete loss of ATPase activity caused by the MLH1 E37K variant cannot be explained solely by loss of the catalytic carboxylate. However, we note that the corresponding aqMutL E32K variant analyzed in this study also exhibited essentially no detectable ATPase activity, indicating that this phenotype is not unique to MLH1. It can be thought that the severe defect of the lysine variants arises not merely from loss of the acidic side chain but from charge reversal. We have clarified this point in the Results and Discussion sections:

“In contrast, the E37K mutation in the MLH1 NTD completely abolished the ATPase activity under our assay conditions (Fig. 5B and Table 1) unlike the corresponding glutamine substitutions, which retained substantial residual ATPase activity in aqMutL, aqGyrB, and PMS2 NTDs. A similar complete loss of ATPase activity was also observed for the E32K mutant form of the aqMutL NTD. These observations suggest that the severe defect caused by the lysine substitution cannot be attributed simply to loss of the catalytic carboxylate. Instead, introduction of a positively charged side chain (charge reversal) is likely to perturb the local electrostatic environment. Structural characterization of the MLH1 E37K and aqMutL E32K mutant forms will be required to clarify the molecular basis of this severe functional defect.” (p. 6 line 281-289 in the revised manuscript)

(3) The authors do not examine ATP binding in the E32 mutants of aqMutL NTD and the D51 mutants of aqGyrB, or AMPPNP binding of the NLH1 and PMS2 mutants. Hence, the relative contributions of the acidic residues to ATP binding and hydrolysis remain partially unclear.

We performed additional ATP-binding experiments using the aqMutL NTD E32A and aqGyrB NTD D51A mutant forms. Both mutant forms exhibited ATP-binding activities comparable to those of the corresponding wildtype forms. These results support our conclusion that the second acidic residue primarily contributes to ATP hydrolysis rather than ATP binding, whereas the first acidic residue plays dual roles in ATP binding and catalysis.

Although we agree that nucleotide-binding analyses of the MLH1 and PMS2 variants would be informative, these experiments were not feasible because the equilibrium dialysis assay requires high protein concentrations, which we were unable to obtain for the recombinant human MLH1 and PMS2 N-terminal domains.

We have incorporated these new data into the Results and Discussion section:

“In contrast to the E29A mutant form of the aqMutL NTD, the E32A mutant form exhibited ATP binding ability comparable to that of the wildtype form (Supplementary Fig. S1A), indicating that Glu32 does not contribute to ATP binding.” (p. 3 line 143-145 in the revised manuscript)

“The D51A mutant form of the aqGyrB NTD retained ATP binding ability comparable to that of the wildtype form, indicating that Asp51 is not required for nucleotide binding (Supplementary Fig. S1B).” (p. 4 line 183-185 in the revised manuscript)

(4) The ATPase assays for PMS2 and MLH1 (Figure 7 and Table 1) were performed with purification/solubility tags still present. Hence, it cannot be ruled out that these tags influence the measured activities.

We thank the reviewer for raising this important point. We agree that the possible influence of the purification/solubility tags on the absolute ATPase activities of the PMS2 and MLH1 NTDs cannot be completely excluded. However, the wild-type and mutant forms for each homolog were analyzed using identical constructs under the same experimental conditions. Therefore, the affinity/solubility tags are unlikely to affect the relative comparisons of the mutational effects. Furthermore, because the affinity tags are located at the N terminus and are distant from the ATPase active site, they are unlikely to directly perturb the catalytic center.

(5) The authors suggest that the two-acidic-residue mechanism proposed in this study could be shared among several GHKL ATPase families, yet they also state that the hydrogen-bonding network was not observed in MutL and MORC family proteins. This raises doubt about how conserved the mechanism is, e.g., in MutL and MORC proteins.

We thank the reviewer for this insightful comment. Our proposed mechanism is based on the cooperative catalytic roles of the two conserved acidic residues, namely the involvement of the first acidic residue in ATP binding and nucleophilic water positioning and the role of the second acidic residue in proton abstraction. In contrast, the Glu48–Gln340 hydrogen-bonding interaction described in aqGyrB was proposed only as a structural feature that may modulate the contribution of the first acidic residue to ATP binding. It is not an essential component of the catalytic mechanism proposed in this study. Therefore, the absence of this particular hydrogen-bonding network in the currently available structures of MutL and MORC proteins does not argue against conservation of the catalytic mechanism itself.

Recommendations for the authors:

Reviewing Editor Comments:

One of the structures (Crystal Structure of the E48A variant) has relatively poor statistics in the PDB validation report. Please improve this structure.

We performed additional refinement of the E48A crystal structure. This resulted in a clear improvement in the overall model quality, with the Ramachandran favored residues increasing from 93.4% to 95.4%, the percentage of side-chain outliers decreasing from 6.1% to 1.4%. The refined structural model has been used throughout the revised manuscript, and the updated refinement statistics are provided in Table 2.

Reviewer #1 (Recommendations for the authors):

Please show conventional density maps (e.g., sigmaA weighted 2fo-fc maps).

This comment is closely related to Comment (2) in the Public Review by the Reviewer #1. In response, we have added Supplementary Fig. S2, which presents conventional σA-weighted 2Fo–Fc electron density maps around the catalytic acidic residues in the wild-type and mutant aqGyrB structures.

Reviewer #2 (Recommendations for the authors):

(1) Regarding the analysis of clinical variants, it would be informative to note that the second allele is lost before tumor growth in Lynch syndrome.

“Therefore, these variants might contribute to the development of Lynch syndrome by weakening the ATPase-driven regulatory functions of MutL.” (p. 6 line 280-281 in the original manuscript) has been changed to:

“In individuals carrying these germline variants, subsequent loss or inactivation of the remaining wildtype allele would leave only the ATPase-defective MutL protein, thereby compromising mismatch repair and promoting tumorigenesis.” (p. 6 line 296-298 in the revised manuscript)

(2) P. 4, in the paragraph "Conserved roles of two acidic residues of aqGyrB in ATP hydrolysis", the E48Q mutant retains approximately one third of the WT activity, not one quarter as stated in the text (Table 1). Additionally, later in the article, the D51 mutant is reported to retain approximately one-sixth (~17%) of the WT activity, rather than ~25% as written.

We thank the reviewer for carefully identifying these inconsistencies. The text has been corrected to accurately reflect the data presented in Table 1: “…one third of the wildtype activity” (p. 4 line 180) and “…retaining ~16%...” (p. 4 line 186 in the revised manuscript)

(3) P. 6, lines 275-276, this sentence should be rephrased for clarity, as the authors note at the end of page 5 that not all members of the GHKL ATPase family possess this second acidic residue.

“…this second acidic residue plays a conserved and functionally significant role in ATP hydrolysis across the GHKL ATPase family.” in the original manuscript has been changed to:

“…this second acidic residue plays a conserved and functionally significant role in ATP hydrolysis among some members of the GHKL ATPase family.” (p. 6 line 292 in the revised manuscript)

(4) P. 9, in the "Data Accessibility Statement", the PDB code 23UY is missing. This entry corresponds to the crystal structure of the D51A mutant of aqGyrB NTD and should be included.

The Data Accessibility Statement has been revised to include the code 23UY. (p. 9 line 454 in the revised manuscript)

(5) P. 14, the table should be labelled "Table 2. Data collection and refinement statistics for the aqGyrB NTDs", rather than "Supplementary Table 2", to ensure consistency with how it is cited in the main text.

The table title has been corrected from "Supplementary Table 2" to "Table 2”. (p. 4 line 198 in the revised manuscript)

(6) It is difficult to determine from the figures whether the magnesium ion is positioned equivalently in aqMutL and aqGyrB. Did the authors observe any differences in ion positioning?

To facilitate direct comparison of the catalytic Mg2+ ion between the aqMutL and aqGyrB NTDs, we have added Supplementary Fig. S3, which shows a structural superimposition of the ATPase active sites of the two proteins:

“Structural superposition of the aqGyrB NTD and aqMutL NTD revealed that the catalytic Mg2+ ion occupies essentially the same position in the two ATPase active sites (Supplementary Fig. S3), indicating that the metal-binding geometry is highly conserved, where the Mg2+ ion is coordinated by the side chain of the conserved Asn, AMPPNP, and surrounding water molecules. Neither Glu48 of aqMutL nor Asp51 of aqGyrB directly coordinated the Mg2+ ion.” (p. 5 line 201-205 in the revised manuscript)

(7) The authors should discuss the interaction between the aqGyrB NTD, Mg2+, and ATP during the binding step. In the case of the E48A mutant, where ATP binding is lost, does E48 directly establish contacts with Mg2+, or is another residue involved (with conformational changes preventing this interaction)?

Our structural analyses indicate that Glu48 does not directly coordinate the catalytic Mg2+ ion. Instead, as shown in Supplementary Fig. S3, the Mg2+ ion is coordinated by the side chain of Asn52, AMPPNP, and surrounding water molecules. We have clarified this point in the Results and Discussion sections:

“Structural superposition of the aqGyrB NTD and aqMutL NTD revealed that the catalytic Mg2+ ion occupies essentially the same position in the two ATPase active sites (Supplementary Fig. S3), indicating that the metal-binding geometry is highly conserved, where the Mg2+ ion is coordinated by the side chain of Asn52, AMPPNP, and surrounding water molecules. Neither Glu48 nor Asp51 directly coordinated the Mg2+ ion.” (p. 5 line 201-205 in the revised manuscript)

(8) Figures 1 and 3: Use ribbon representation and no shadows, at least for the inset panels, to enhance clarity and interpretability.

We have revised Figures 1 and 3 by displaying the protein structures in ribbon representation and removing shadows from the inset panels.

(9) Combine Figures 1 and 2, and combine Figures 3 and 4.

Following the reviewer's recommendation, we have combined the original Figures 1 and 2 into a single figure and the original Figures 3 and 4 into another single figure.

(10) Figure 5: Zoom in further and remove shadows. The current panels are not very effective in highlighting how ATP is bound by the different protein variants.

Figure 5 has been revised by increasing the magnification of the ATP-binding sites and removing shadows from the structural renderings.

(11) Figure 8. Add a scale bar to show evolutionary distance.

We thank the reviewer for this helpful suggestion. To provide information on evolutionary distances while preserving the clarity of the main figure, we have added a new Supplementary Fig. S5 showing the same phylogenetic tree with branch lengths proportional to the inferred evolutionary distances and an evolutionary distance scale bar. Figure 6 has been retained in its simplified form with equal branch lengths to facilitate visualization of the ancestral-state reconstruction, and we have clarified this distinction in the Materials and Methods section:

“For visualization purposes, branch lengths were not scaled and were displayed with equal lengths in Fig. 6. The corresponding phylogeny with branch lengths proportional to the inferred evolutionary distances is provided in Supplementary Fig. S5.” (p. 9 line 429-432 in the revised manuscript)

Version 2

Abstract

GHKL ATPases share a unique Bergerat ATP-binding fold and regulate diverse biological processes through ATP-dependent conformational changes. An early step of ATP hydrolysis in this family has been attributed to a single highly conserved glutamate residue proposed to function as the general base. However, mutations of this residue impair both the ATPase activity and ATP binding, complicating interpretation of its catalytic role. Re-examination of the high-resolution crystal structures revealed a second conserved acidic residue positioned within a hydrogen-bonding distance from the nucleophilic water molecule. Using Aquifex aeolicus MutL and GyrB as model enzymes, we combined systematic mutagenesis, ATPase and ATP-binding assays, and X-ray crystallography to dissect the roles of these residues. We show that alignment of the nucleophilic water can be maintained as long as the conserved glutamate retains hydrogen-bonding capability, whereas efficient ATP hydrolysis requires proton- accepting capacity at least one of the two acidic residues. These results indicate that the conserved glutamate primarily governs positioning of the nucleophilic water, while activation of this water for catalysis is achieved through cooperative general base function of the two acidic residues. Extending this framework to human MutL homologs, PMS2 and MLH1, we showed that clinically reported variants of uncertain significance in these DNA mismatch repair proteins substantially reduced the ATPase activity, indicating functional impairment. Together, our findings refine the catalytic mechanism of GHKL ATPases and provide a structural and functional framework for interpreting disease-associated variants in GHKL ATPases. Phylogenetic and ancestral state analysis further indicated that the second acidic residue was likely to be present in the common ancestor of major GHKL ATPase lineages but was later modified in a branch including Hsp90, suggesting evolutionary remodeling of the catalytic mechanism in the branch.

Introduction

The GHKL ATPase family comprises a wide range of proteins involved in diverse cellular processes, including DNA replication, DNA repair, protein homeostasis, chromatin regulation, and signal transduction 1. The GHKL ATPase family includes the following representative members: DNA gyrase B (GyrB), the ATPase subunit of type II topoisomerase that introduces negative supercoils into DNA 2,3; the molecular chaperone Hsp90, which regulates the folding and stability of numerous client proteins 4; histidine kinases of two-component regulatory systems, which function in environmental sensing and signal transduction 5; the DNA mismatch repair protein MutL, which maintains genome stability by correcting replication errors 6,7; and MORC family proteins, which contribute to chromatin organization and epigenetic regulation 8,9. Despite their functional diversity, these proteins share a conserved Bergerat ATP-binding motif and are thought to employ a common mechanism for ATP binding and hydrolysis 1,10–13.

For GHKL ATPases, ATPase activity is essential for protein function. In many cases, ATP binding and hydrolysis drive large-scale conformational changes that are coupled to biological activity. The ATP-bound and ADP-bound states often correspond to distinct oligomeric or domain arrangements, and cycling between these states enables GHKL ATPases to act as molecular switches or motors that coordinate complex cellular processes 14–16. A representative example of such functional coupling is provided by MutL, a key component of the DNA mismatch repair pathway 17–19. MutL consists of an N-terminal ATPase domain and a C-terminal nuclease domain, and ATP binding induces dimerization of the N-terminal domains, leading to rearrangement of the overall architecture and engagement of downstream repair factors 14,20–22. Subsequent ATP hydrolysis, stimulated by interacting proteins and DNA substrates, induces further conformational changes that activate its nuclease function and enable strand incision during mismatch repair 7,23–25. Thus, in MutL, the ATP binding and hydrolysis cycle directly controls large-scale structural transitions that are essential for its biological activity. A detailed understanding of the ATP hydrolysis mechanism is therefore crucial not only for elucidating the molecular basis of fundamental biological phenomena but also for clarifying disease mechanisms associated with dysfunction of GHKL ATPases.

In human, impaired function of MutL proteins compromises the DNA mismatch repair pathway, leading to reduced fidelity of DNA replication. Such defects are a major cause of Lynch syndrome, one of the most common hereditary cancer predisposition syndromes 26–30. Cancers associated with Lynch syndrome exhibit high responsiveness to immune checkpoint inhibitors 31,32, and genetic testing of MutL genes is recommended for diagnosis. However, clinical interpretation is often hampered by the presence of variants of uncertain significance. This limitation largely reflects the insufficient understanding of MutL function at the amino acid level. Variants of uncertain significance have also been identified in the ATPase active site of human MutL homologs, highlighting the importance of elucidating the ATP hydrolysis mechanism of MutL for establishing a molecular basis for genetic diagnosis of Lynch syndrome.

Among GHKL ATPases, GyrB, MutL, and MORC are thought to share particularly similar ATPase mechanisms 11. Within the Bergerat ATP-binding motif, a highly conserved glutamate residue, Glu29 in Aquifex aeolicus MutL (aqMutL), has been proposed to function as a general base that activates the nucleophilic water molecule for ATP hydrolysis 33,34. This proposal is based on observations that substitution of alanine or lysine for this glutamate abolishes the ATPase activity. In addition, a conserved lysine residue, Lys79, in aqMutL, has been suggested to act as a general acid, donating a proton to the reaction intermediate and thereby facilitating cleavage of the γ-phosphate 11. Other conserved basic residues have also been proposed to contribute as Lewis acids that stabilize negatively-charged reaction intermediates 11,34. Collectively, these residues have been proposed to form the core catalytic machinery of GHKL ATPases.

Previously, we determined high-resolution crystal structures of the ATPase domain of aqMutL in complex with ATP analogs, in which a water molecule presumed to be the nucleophile for ATP hydrolysis was clearly observed at a hydrogen-bonding distance to the conserved glutamate Glu29 35. Upon careful re-examination of these structures, we noticed that this same water molecule is also positioned within a hydrogen-bonding distance to a second conserved acidic residue, Glu32 (Fig. 1A). Thus, the nucleophilic water is not exclusively coordinated by Glu29, but simultaneously interacts with two acidic residues.

Effects of mutations on the overall structure and the ATPase activity of the aqMutL NTD.

(A) Crystal structure of the aqMutL N-terminal domain (NTD) in complex with the non-hydrolyzable ATP analog AMPPCP. The boxed region indicates the ATP-binding site and is shown enlarged on the right. AMPPCP and the side chains of Glu29 and Glu32 are shown as stick models. Oxygen atoms are shown in red, nitrogen atoms in blue, and phosphorus atoms in orange. The magnesium ion and the water molecule putatively acting as the nucleophile are depicted as green and red spheres, respectively. Hydrogen bonds are indicated by dashed lines, with distances shown. The blue mesh represents an omit Fo–Fc electron density map contoured at 3σ for AMPPNP, the magnesium ion, and the water molecule. (B) Sequence alignment of representative GHKL ATPases. The Bergerat ATP-binding fold is underlined, and the conserved acidic residues corresponding to aqMutL Glu29 and Glu32 are marked with asterisks. Amino acid sequence alignment was performed by CLUSTAL W program 57 and visualized by ESPript3 58. (C) Far-UV CD spectra of the wildtype and mutant forms of the aqMutL NTD were collected and averaged over 10 consecutive scans. (D, E) ATPase activity of the aqMutL NTD was measured under steady-state conditions using a colorimetric assay that detects the release of inorganic phosphate. Apparent rate constants (kapp) were calculated from the measured phosphate concentrations and plotted as a function of ATP concentration. Kinetic parameters were determined by fitting Michaelis–Menten equation to the data. Each data point represents the mean of three independent experiments, with error bars indicating standard deviation. Theoretical curves calculated from the fitted kinetic parameters are overlaid. Data for 0–1 mM substrate concentrations are shown to compare the wildtype from with mutant forms (D). Activities for higher substrate concentrations were measured for some of the mutant forms (E).

Although earlier biochemical studies, including those from our group, demonstrated that replacement of Glu29 with alanine abolishes the ATPase activity, such mutations also severely impair ATP binding 11,33, complicating interpretation of the specific catalytic role of this residue. In this context, the potential contribution of Glu32 to ATP hydrolysis might have been overlooked. Supporting this notion, the residue corresponding to Glu32 is conserved not only in MutL but also in other GHKL ATPases, including GyrB and MORC proteins (Fig. 1B). Furthermore, replacements of the corresponding Glu residue with Lys in Escherichia coli MutL and in human MutL homolog 1 (MLH1) have been reported to abolish DNA mismatch repair activity in cell-based or nuclear extract- based assays 36,37.

In this study, we investigated the functional role of Glu32 in the ATP hydrolysis mechanism of MutL and GyrB using a combination of enzymatic and X-ray crystallographic analyses. Our results provide new insights into the catalytic mechanism of GHKL ATPases and refine the current model of ATP hydrolysis for this protein family. By clarifying the molecular basis of ATP hydrolysis in GHKL ATPases, this work also provides a mechanistic framework for interpreting the functional consequences of some clinically observed variants, thereby contributing to the foundation for assessing the impact of variants of uncertain significance.

Results and discussion

Mutational analysis of Glu29 and Glu32 in the ATPase domain of aqMutL

To examine the contribution of Glu32 to the ATPase activity of the aqMutL NTD and to directly compare its role with that of Glu29, we introduced a series of substitutions at both positions. To exclude the possibility that the observed effects on the ATPase activity arose from global structural perturbations, circular dichroism (CD) spectra were measured for all mutant forms (Fig. 1C). Among these, only the E32G mutant form exhibited a marked decrease in ellipticity, suggesting that this substitution disrupts the secondary structure of the protein. In contrast, all other mutant forms displayed CD spectra indistinguishable from that of the wildtype form, indicating that these substitutions do not substantially perturb the overall structure of the aqMutL NTD.

We first analyzed the effects of mutations at Glu29 on the ATPase activity of the aqMutL NTD. The mutation to Gln would selectively impair the proton association/dissociation capability of the Glu side chain while preserving its size and hydrogen-bonding potential. Moreover, because mutations to Val, Lys, and Gly at this position have been reported as variants of uncertain significance in human MutL, these mutations were also analyzed. Consistent with previous reports, all Glu29 mutant forms exhibited severe reduction in the ATPase activity (Fig. 1D and E, and Table 1). In particular, the E29A, E29V, E29K, and E29G mutant forms showed activities below the detection limit, indicating that Glu29 plays a central role in ATP hydrolysis. However, as described above, previous studies also demonstrated that mutations at Glu29 impair ATP binding 11,33,34, indicating that the loss of the ATPase activity of these mutant forms cannot be unambiguously attributed to disruption of a catalytic base function.

To further dissect the role of Glu29, we focused on the E29Q mutant form. Using equilibrium dialysis, we confirmed that the E29Q mutant form binds ATP with an affinity comparable to that of the wildtype protein (Supplementary Fig. S1A). Because this substitution does not impair ATP binding, it should enable a more direct assessment of the proposed role of Glu29 as a general base in ATP hydrolysis. The E29Q mutant form retained measurable ATPase activity, although its kcat value was approximately tenfold smaller than that of the wildtype enzyme (Fig. 1D and E, and Table 1). The persistence of substantial catalytic activity in the absence of the carboxylate group at the position 29 suggests that a residue other than Glu29 can participate in proton abstraction from the nucleophilic water during ATP hydrolysis.

We next examined the effects of mutations at Glu32. In contrast to the E29A mutant form of the aqMutL NTD, the E32A mutant form exhibited ATP binding ability comparable to that of the wildtype form (Supplementary Fig. S1A), indicating that Glu32 does not contribute to ATP binding. Mutations of Glu32 resulted in reduced ATPase activity relative to the wildtype form, supporting a role for this residue in catalysis. However, with the exception of the E32K and the structurally perturbed E32G mutant forms, all Glu32 mutant forms retained detectable ATPase activity (Fig. 1D and E, and Table 1), indicating that loss of Glu32 alone is insufficient to abolish ATP hydrolysis. Simultaneous substitutions of glutamine for the two residues in the E29Q/E32Q double mutant form led to a complete loss of the ATPase activity. Taken together, these results indicate that Glu29 and Glu32 do not exert equivalent functions but act cooperatively. The presence of at least one carboxylate group at either position, 29 or 32, appears to be required for ATP hydrolysis.

Conserved roles of two acidic residues of aqGyrB in ATP hydrolysis

To examine whether the cooperative roles of Glu29 and Glu32 observed in aqMutL are conserved among other GHKL ATPases, we next analyzed the NTD of A. aeolicus GyrB (aqGyrB). A glutamate residue corresponding to aqMutL Glu29 is also conserved in bacterial GyrB and has been proposed to function as a general base in ATP hydrolysis 38. Although crystal structure of the E. coli GyrB NTD has been reported 39, the position of the putative nucleophilic water has not been clearly defined. Therefore, we selected the GyrB NTD from the hyperthermophilic bacterium A. aeolicus as a model system, anticipating that it would allow high-resolution structural analysis and clearer visualization of a catalytic water molecule.

First, we determined the crystal structure of this protein in complex with AMPPNP at 1.65 Å resolution (Table 2 and Fig. 2A). The overall structure was highly similar to that of the E. coli GyrB NTD, with a Cα root mean square deviation of 1.42 Å. As observed for the E. coli GyrB NTD, AMPPNP was accommodated by the Bergerat ATP-binding fold, and AMPPNP binding seemed to induce conformational changes that promoted dimer formation through the newly formed interface. Glu48 and Asp51 of aqGyrB correspond to Glu29 and Glu32 of aqMutL, respectively. In the crystal structure, these residues were found to tightly coordinate a water molecule that is likely to act as the nucleophilic water during ATP hydrolysis (Fig. 2A). Based on these structural observations, we introduced Ala or Gln/Asn substitutions for Glu48 and Asp51. The CD spectra of the mutant forms were nearly identical to that of the wildtype form (Fig. 2B), suggesting that these substitutions did not disrupt the overall structure of the aqGyrB NTD.

Crystal structure of the aqGyrB NTD and effects of mutations on the overall structure and the ATPase activity of the aqGyrB NTD.

(A) Structure of the AMPPNP-bound dimer of the wildtype aqGyrB NTD. The boxed region highlights the ATP-binding site and is enlarged in the left panel. The same region is further enlarged and rotated 60° to provide an alternative view of the active site geometry. AMPPNP and the side chains of Glu48 and Asp51 are shown as stick models. Oxygen atoms are shown in red, nitrogen atoms in blue, and phosphorus atoms in orange. The magnesium ion and the water molecule presumed to act as the nucleophile are depicted as green and red spheres, respectively. Hydrogen bonds are indicated by dashed lines, with distances shown. The blue mesh represents an omit Fo–Fc electron density map contoured at 3σ for AMPPNP, the magnesium ion, and the water molecule. (B) Far-UV CD spectra of the wildtype and mutant forms of the aqGyrB NTD were recorded and averaged over 10 accumulations. (C) ATPase activity of the aqGyrB NTD was measured under steady-state conditions as described in Fig. 1. The kapp values were plotted as a function of ATP concentration. Michaelis–Menten equation was fitted to the data. Data points represent the mean ± standard deviations from three independent experiments.

Equilibrium dialysis showed that the E48A mutant form failed to bind ATP, whereas the E48Q mutant form displayed ATP-binding affinity comparable to that of the wildtype protein (Supplementary Fig. S1B). The E48Q mutation therefore allowed us to evaluate the catalytic contribution of Glu48 independently of ATP binding. ATPase activity assays showed that the E48A mutant form completely lacked the activity, while the E48Q mutant form retained approximately one third of the wildtype activity (Fig. 2C and Table 1). The persistence of substantial catalytic activity in the E48Q mutant form suggests that another residue partially compensates for the loss of the Glu48 carboxylate in the proton abstraction.

The D51A mutant form of the aqGyrB NTD retained ATP binding ability comparable to that of the wildtype form, indicating that Asp51 is not required for nucleotide binding (Supplementary Fig. S1B). Mutations of Asp51 also impaired ATP hydrolysis, with the D51A and D51N mutant forms retaining ∼16% and ∼50% of the wildtype activity, respectively (Fig. 2C and Table 1). Notably, the E48Q/D51N double mutant form exhibited no detectable ATPase activity, resembling the effect of the corresponding double mutation in the aqMutL NTD. Taken together, these results indicate that both Glu48 and Asp51 contribute to the catalytic mechanism of ATP hydrolysis in aqGyrB and that at least one of these residues must retain a carboxylate group with the proton-accepting capability to sustain the enzymatic activity.

Crystal structures of aqGyrB mutant forms and the role of Glu48 in ATP binding

In order to relate the ATPase activity profiles of the mutant forms to their structural features, we determined X-ray crystal structures of selected mutant forms using the aqGyrB NTD as a model system, which allows relatively high-resolution structural analyses. The AMPPNP-bound structures of the E48A, E48Q, D51A, and D51N mutant forms were determined at resolutions of 2.66, 1.70, 1.63, and 2.05 Å, respectively (Table 2). To show that the introduced substitutions were unambiguously supported by the crystallographic data, electron density maps around residues 48 and 51 are shown in Supplementary Fig. S2.

Structural superposition of the aqGyrB NTD and aqMutL NTD revealed that the catalytic Mg2+ ion occupies essentially the same position in the two ATPase active sites (Supplementary Fig. S3), indicating that the metal-binding geometry is highly conserved, where the Mg2+ ion is coordinated by the side chain of Asn52, AMPPNP, and surrounding water molecules. Neither Glu48 nor Asp51 directly coordinated the Mg2+ ion.

Since the conserved general base Glu (i.e. Glu48 of aqGyrB) does not directly interact with ATP, it has been unclear why mutations at the Glu residue abolish the ATP-binding ability of GHKL ATPases. To clarify this, we compared the crystal structures of the wildtype and mutant forms of the aqGyrB NTD (Fig. 3). In GHKL ATPases, ATP binding is tightly coupled to conformational rearrangements, including formation of an ATP lid (Fig. 3A), which acts to secure the bound nucleotide by enclosing it within the active site 3,21. Structural analysis of the wildtype aqGyrB NTD revealed that the side chain of Glu48 forms a hydrogen bond with Gln340, thereby stabilizing a loop containing Gln340 in a conformation capable of ATP binding (Fig. 3B). In this arrangement, Gln340 in turn forms a hydrogen bond with His122, a residue located within the ATP lid, while Lys342 in the same loop is positioned to interact with the γ-phosphate of ATP. These coordinated interactions appear to promote the ATP-induced conformational transition required for stable nucleotide binding.

Structural basis for the loss and preservation of ATP binding in aqGyrB NTD mutant forms.

The wildtype and all mutant forms were crystallized in the presence of AMPPNP; however, the E48A mutant form did not bind AMPPNP, whereas the E48Q, D51A, and D51N mutant forms were observed in complex with AMPPNP. (A) Overall structures of the wildtype and E48A mutant aqGyrB NTD monomers. The ATPase active site is indicated by a boxed region. The ATP-lid (residues 105–124) in the wildtype form and the corresponding region in the E48A mutant form are shown in magenta and olive, respectively. (B–F) Enlarged views of the ATPase active site in the wildtype (B), E48A (C), E48Q (D), D51A (E), and D51N (F) forms. Side chains of the residues implicated in the ATP-induced conformational changes (e.g. His122, Gln340, and Lys342), as well as residues 48 and 51, are shown as stick models. AMPPNP is depicted as sticks, and the Mg2+ ion is shown as a green sphere. Hydrogen bonds and ionic interactions are indicated by dashed lines.

In contrast, structural analysis of the E48A mutant form demonstrated that disruption of the interaction with Gln340 abolishes this network of hydrogen bonds, preventing the associated conformational rearrangements and thereby impairing ATP binding (Fig. 3C). These findings indicate that the interaction network emanating from Glu48 is indispensable for the conformational changes associated with the ATP lid formation. Interestingly, in the E48Q mutant form, the hydrogen bond with Gln340 is preserved, allowing the conformational transition to occur and enabling ATP binding (Fig. 3D). Consistently, in the D51A and D51N mutant forms, the interaction mediated by Glu48 remains intact, and the structural rearrangements required for ATP binding are maintained (Fig. 3E and F).

These structural observations extend previous models of GyrB ATPase function by linking the conserved catalytic glutamate to ATP lid stabilization and nucleotide-dependent conformational rearrangement. The E48A aqGyrB NTD structure, in particular, provides direct structural evidence that disruption of this Glu-centered interaction network is sufficient to prevent ATP-induced active- site assembly. This hydrogen-bonding network emanating from the general base Glu is conserved in GyrBs from other organisms 39,40 (Supplementary Fig. S4), while it is not observed in other GHKL ATPases, including MutL 22,34,35,41 and MORC family proteins 9,42,43. Therefore, the loss of ATP- binding ability observed upon the mutation of Glu29 to Ala in MutL 11 is likely to be explained by a different mechanism. This difference highlights mechanistic diversity within the GHKL ATPase family, despite conservation of the two catalytic acidic residues.

Nucleophilic water molecule in the active site of mutant forms of the aqGyrB NTD

Having established the structural basis for the altered ATP-binding properties of the mutant forms, we next analyzed how these substitutions influence the positioning of the nucleophilic water molecule. Structural comparisons of the catalytic sites of the four mutant forms are shown in Fig. 4. In the E48A mutant form (Fig. 4A), the nucleophilic water molecule was not observed in the catalytic site. In contrast, in the E48Q mutant form, the nucleophilic water molecule was positioned at essentially the same location as in the wildtype form, coordinated by Gln48 and Asp51, and bound AMPPNP was observed (Fig. 4B). This structural arrangement is consistent with the biochemical data described above, in which the E48Q mutant form retained substantial ATP-binding and ATPase activities.

Structures surrounding the conserved acidic residues: (A) E48A, (B) E48Q, (C) D51A, and (D) D51N mutant forms of the aqGyrB NTD.

AMPPNP and the side chains at the positions 48 and 51 are shown as stick models. Oxygen atoms are shown in red, nitrogen atoms in blue, and phosphorus atoms in orange. The magnesium ion is shown as a green sphere, and the water molecule proposed to act as the nucleophilic water is shown as a red sphere. Blue mesh represents Fo–Fc omit electron density maps contoured at 3σ, calculated after omission of AMPPNP, the magnesium ion, and the nucleophilic water molecule. Hydrogen bonds are indicated by dashed lines, with the corresponding distances shown.

Unlike the E48A mutant form, the D51A mutant form bound AMPPNP (Fig. 4C), and the nucleophilic water molecule was coordinated only by Glu48. This asymmetric coordination suggests that Glu48 plays a dominant role in positioning the nucleophilic water molecule. In the D51N mutant form (Fig. 4D), which lacks the proton-transfer capability of Asp51 while retaining its hydrogen- bonding capacity, the nucleophilic water was coordinated by both Glu48 and Asn51. These structural features are consistent with the relative ATPase activities observed for the mutant forms of these two acidic residues (Fig. 2C).

Taken together, these results indicate that Glu48 and Asp51 make distinct yet cooperative contributions to ATP hydrolysis. Glu48 appears to play a dominant role in aligning the nucleophilic water molecule and activating this water molecule through proton abstraction as a general base. In contrast, Asp51 seems to contribute less to the precise positioning of the nucleophilic water but participates cooperatively with Glu48 in the proton abstraction, thereby supporting efficient ATP hydrolysis. Notably, an acidic residue corresponding to Asp51 of aqGyrB is conserved among several GHKL ATPases, including MutL, GyrB, and MORC family proteins (Fig. 1B), suggesting that the two-acidic-residue mechanism proposed here may be shared among these subgroups. The second acidic residue is not conserved in other members of the GHKL family, such as Hsp90 and mitochondrial histidine kinases (Fig. 1B), where a single acidic residue is thought to fully function as the general base responsible for water activation. A similar divergence appears to apply to the previously-reported general acid Lys residue 11, further supporting the notion that GHKL ATPases can be divided into at least two mechanistically distinct subclasses. This mechanistic bifurcation provides insights into the molecular evolution of the GHKL ATPase family.

Functional and clinical implications of the second acidic residue in human MutL homologs

Our analyses of the aqMutL NTD and aqGyrB NTD indicate that, in addition to the Glu residue traditionally considered to function as the general base, an additional conserved acidic residue (Glu or Asp) also participates in the general base catalysis. In human MutL homologs, this residue corresponds to Glu44 in Postmeiotic Segregation Increased 2 (PMS2) and Glu37 in MLH1. Notably, variants of uncertain clinical significance, including E44V (ClinVar variation ID: 232732), E44G (230139), and E44Q (186069) in PMS2, as well as E37G (230139) and E37K (89640) in MLH1, have been reported at these positions.

To examine whether these residues similarly contribute to the ATPase activity in human MutL homologs, we introduced the corresponding mutations into the NTDs of human PMS2 and MLH1 and measured their ATPase activities. The recombinant E44G human PMS2 NTD and E37G human MLH1 NTD could not be expressed successfully, which is consistent with our finding that the corresponding mutation in the aqMutL NTD disrupted the structural integrity of the protein (Fig. 1C). Both the E44V and E44Q mutant forms of the PMS2 NTD exhibited a marked reduction in the ATPase activity, retaining approximately one-sixth of the wildtype activity (Fig. 5A and Table 1). In contrast, the E37K mutation in the MLH1 NTD completely abolished the ATPase activity under our assay conditions (Fig. 5B and Table 1) unlike the corresponding glutamine substitutions, which retained substantial residual ATPase activity in the aqMutL, aqGyrB, and PMS2 NTDs. A similar complete loss of ATPase activity was also observed for the E32K mutant form of the aqMutL NTD. These observations suggest that the severe defect caused by the lysine substitution cannot be attributed simply to loss of the catalytic carboxylate. Instead, introduction of a positively charged side chain (charge reversal) is likely to perturb the local electrostatic environment. Structural characterization of the MLH1 E37K and aqMutL E32K mutant forms will be required to clarify the molecular basis of this severe functional defect.

ATPase activities of human MutL homolog NTDs.

(A) ATPase activity of mutant forms of the human PMS2 NTD. (B) ATPase activity of a mutant form of the human MLH1 NTD. ATPase activities were measured using the same procedures as described for Fig. 1 and Fig. 2. Apparent rate constants (kapp) were plotted as a function of substrate concentration. Data points represent the mean values from three independent measurements, with error bars indicating standard deviations. Solid lines represent the theoretical Michaelis-Menten curves.

These results are in good agreement with our observations for the aqMutL NTD and aqGyrB NTD, further supporting the notion that this second acidic residue plays a conserved and functionally significant role in ATP hydrolysis among some members of the GHKL ATPase family. Previous studies demonstrated that even a several-fold reduction in the MutL ATPase activity could severely compromise mismatch repair efficiency in vivo 11. In this context, the substantial reduction in the ATPase activity caused by the E44V and E44Q mutations in the PMS2 NTD is likely to impair mismatch repair in human cells. In individuals carrying these germline variants, subsequent loss or inactivation of the remaining wildtype allele would leave only the ATPase-defective MutL protein, thereby compromising mismatch repair and promoting tumorigenesis. Nevertheless, whether these variants indeed compromise mismatch repair activity in vivo remains to be directly tested. Functional validation using cell-based or nuclear extract-based mismatch repair assays, such as the 6-thioguanine sensitivity assay 44 or CIMRA assay 36, will be required to establish a definitive link between the reduced ATPase activity and mismatch repair deficiency in human cells.

Although we focused here on pathogenic variants in the MutL homologs MLH1 and PMS2, extending similar structural and biochemical analyses to disease-associated variants in other human GHKL ATPases will be important for evaluating the generality and clinical relevance of the conserved catalytic mechanism proposed in this study.

Phylogenetic analysis of GHKL ATPases

The newly-identified functionally-important acidic residue is not uniformly conserved across the GHKL superfamily. We therefore used this position as an evolutionary marker to investigate patterns of molecular divergence within the GHKL ATPase domain. In histidine kinases, even the canonical catalytic glutamate is typically replaced, and the ATPase-associated catalytic machinery is not maintained 1,45,46. Therefore, phylogenetic reconstruction was restricted to other GHKL ATPase family members MutL, GyrB, MORC, and Hsp90.

Phylogenetic analysis based on the trimmed ATPase-domain alignment recovered these major GHKL ATPase lineages (MutL, GyrB, MORC, and Hsp90) as a well-supported cluster, indicating relatively close evolutionary relationships among these groups, which is consistent with a previous study 1 (Supplementary Fig. S5). Ancestral state reconstruction performed on the maximum-likelihood phylogeny strongly supports the presence of an acidic residue at the position corresponding to aqMutL Glu32 or aqGyrB Asp51 in the internal node uniting the MutL, GyrB, MORC, and Hsp90 clades with a posterior probability of 0.983 (Fig. 6). In contrast, the ancestral node of the Hsp90 clade is strongly supported to encode serine at this position with a posterior probability of 1.000. Because no external outgroup was specified, the global polarity of amino acid residue changes across the entire phylogeny cannot be formally determined. Nevertheless, the widespread conservation of the second acidic residue across the MutL, GyrB, and MORC lineages, together with the high posterior probability at the most recent common ancestor of these groups and Hsp90, indicates that this residue was very likely to be present in their ancestor. Its absence within the Hsp90 lineage is therefore most parsimoniously interpreted as lineage-specific loss. In summary, these results suggest that the Hsp90 lineage underwent remodeling of its ATPase catalytic mechanism during its evolutionary divergence from other GHKL ATPase family members.

Maximum-likelihood phylogeny of the ATPase domain of representative GHKL family members.

The state of the amino acid residue at the position corresponding to aqMutL Glu32 or aqGyrB Asp51 is indicated by colored bars (red, acidic residue; blue, non-acidic residue). The phylogenetic tree was visualized and annotated using the Interactive Tree of Life (iTOL) web server56. The tree is displayed in a rectangular layout for clarity and should be interpreted as unrooted. Ancestral state reconstruction was performed on the fixed topology, and posterior probabilities are shown for selected internal nodes. The internal node uniting the MutL/GyrB/MORC/Hsp90 clades strongly supports an acidic residue with the posterior probability (PP) of 0.983, whereas the Hsp90 stem ancestor is strongly supported to encode serine at this position with the posterior probability of 1.000. Species abbreviations used in the phylogenetic tree are as follows: Ce, Caenorhabditis elegans; Sc, Saccharomyces cerevisiae; Dr, Danio rerio; Hs, Homo sapiens; Aa, Aquifex aeolicus; Hp, Helicobacter pylori; Tt, Thermus thermophilus; Ec, Escherichia coli; Mt, Mycobacterium tuberculosis; Bs, Bacillus subtilis; Sm, Streptococcus mutans; Dm, Drosophila melanogaster; Clf, Canis lupus familiaris; Mm, Mus musculus; Hc, Hemicordylus capensis; Pv, Patella vulgata; Tf, Tachysurus fulvidraco; Lk, Lepidochelys kempii; Cc, Castor canadensis; Dn, Dasypus novemcinctus.

Conclusions

In this study, we re-examined the catalytic mechanism of GHKL ATPases using aqMutL and aqGyrB as model systems with mutational, biochemical, and crystallographic studies. Our results reveal that ATP hydrolysis in these enzymes is not governed by a single general base residue, as has been traditionally assumed, but instead relies on the coordinated action of two conserved acidic residues. Structural analyses indicate that Glu48 of aqGyrB plays a dominant role in aligning the nucleophilic water molecule in an optimal geometry for attack on the γ-phosphate of ATP. However, biochemical and structural evidence from single and double mutant forms indicates that both residues contribute to proton abstraction from the nucleophilic water. Thus, general base catalysis in GyrB and MutL appears to be achieved by redundant proton-accepting residues.

Extending this finding to human MutL homologs, we showed that mutations at the corresponding residues in PMS2 and MLH1—reported as variants of uncertain clinical significance—lead to substantial or complete loss of the ATPase activity. Given that even modest reductions in the MutL ATPase activity are known to severely compromise mismatch repair efficiency in vivo, our findings suggest that these variants might be functionally deleterious. Thus, the mechanistic insights obtained in this study provide a structural and biochemical basis for evaluating the pathogenic potential of clinically ambiguous variants in human MutL homologs.

Materials and methods

Construction of expression plasmids

Expression plasmids encoding mutant forms of the aqMutL NTD (residues 1–315), the aqGyrB NTD (residues 1–397), the human PMS2 NTD (residues 1–365), and the human MLH1 NTD (residues 1– 337) were generated by introducing point mutations into pET-11a/aqMutL NTD 47, pET-11a/aqGyrB NTD 11, pCold ProS2/human PMS2 NTD 35, and pET28-MLH1/human MLH1 NTD 11 plasmids using the PrimeSTAR site-directed mutagenesis protocol (Takara). Primer sequences used for mutagenesis are listed in Supplementary Table S1. DNA sequencing confirmed that no unintended mutations were introduced.

Expression and purification of proteins

The wildtype aqMutL NTD, aqGyrB NTD, ProS2-tagged human PMS2 NTD, and His-tagged human MLH1 NTD were overexpressed and purified as described previously 11,35,47. The mutant forms were overexpressed and purified using the same procedures as for the respective wildtype proteins.

CD spectrometry

CD measurements were performed with a spectropolarimeter, model J-720W (Jasco). Measurements of CD spectra were carried out in a solution comprised of 20 mM Tris-HCl (pH 8.0) (eLANT) and 10 μM protein using a 0.1 cm cell at 25°C. The residue molar ellipticity [Θ] was defined as 100 Θobs/(lc), where Θobs is the observed ellipticity, l is the length of the light path in centimeters, and c is the residue molar concentration of the protein.

ATPase assay

ATPase activities of the aqMutL NTD, aqGyrB NTD, ProS2-tagged human PMS2 NTD, and His- tagged human MLH1 NTD were measured as described previously 35. Reactions were carried out using 10 μM protein in a 20 μl reaction mixture containing 50 mM HEPES–KOH (pH 7.5) (eLANT), 100 mM NaCl, 5 mM MgCl₂, and various concentrations of ATP (Roche). For the aqMutL NTD and aqGyrB NTD, reactions were performed at 70°C for 60 min. For the NTDs of the human MutL homologs, reactions were performed at 37°C for 120 min. All experiments were conducted in triplicate. Kinetic parameters were determined by fitting the data to the standard Michaelis–Menten equation using IgorPro.

Equilibrium dialysis

Equilibrium dialysis experiments were performed as described previously 11. Seventy-five microliters of buffer containing 50 mM Tris–HCl (pH 8.0), 100 mM NaCl, 5 mM MgCl₂, and 50 μM AMPPNP (Roche) were loaded into the buffer chamber of a DispoEquilibrium Dialyzer equipped with a 10 kDa molecular weight cut-off membrane (Harvard Bioscience, Inc.). An equal volume of the aqMutL NTD (0–320 μM) or aqGyrB NTD (0–400 μM) in the same buffer without AMPPNP was placed in the opposing sample chamber. The device was gently agitated at room temperature for 24 h to allow equilibration. The absorbance of the solution in the buffer chamber was measured at 260 nm. The concentration of unbound AMPPNP was calculated using a molar extinction coefficient (ε₂₆₀ = 15,400 M⁻¹ cm⁻¹).

Crystallization and structure determination

Protein crystallization was performed by the sitting-drop vapor diffusion method at 20°C. One microliter of the aqGyrB NTDs (14 mg/ml) containing 0.6 mM AMPPNP (Roche) and 1 mM MgCl₂ was mixed with an equal volume of the reservoir solution and equilibrated against 500 μl of reservoir solution. For the wildtype and D51A mutant forms of the aqGyrB NTD, crystals were obtained using a reservoir solution containing 0.2 M ammonium acetate, 0.1 M HEPES (pH 7.5), and 45% (v/v) 2- methyl-2,4-pentanediol. The E48A mutant form was crystallized in 0.05 M cadmium sulfate hydrate, 0.1 M HEPES (pH 7.5), and 20% (w/v) polyethylene glycol 3,350. The E48Q mutant form was crystallized under identical conditions consisting of 30% (w/v) polyethylene glycol 400, 100 mM HEPES–NaOH (pH 7.5), and 200 mM sodium chloride. The D51N mutant form was crystallized in 0.1 M sodium malonate (pH 7.0) and 12% (w/v) polyethylene glycol 3,350. Crystals of the wildtype and E48A mutant forms were soaked in the respective reservoir solutions supplemented with 30% (v/v) glycerol before being flash-cooled in liquid nitrogen. Crystal of the D51N mutant form was transferred to the reservoir solution containing 30% (w/v) polyethylene glycol 3,350 prior to flash- cooling in liquid nitrogen. Crystal of the D51A mutant form was directly flash-cooled in liquid nitrogen without additional cryoprotection.

X-ray diffraction data were collected at beamline BL45XU in SPring-8 (Hyogo, Japan) at a wavelength of 1.000 Å and a temperature of −173°C. Data were automatically collected and processed using the ZOO system 48, in which data processing was assisted by KAMO 49. The linearity of the Wilson plot was confirmed over the resolution range employed. Phases were determined by molecular replacement using Phaser with a search model predicted by AlphaFold2 50. Model building and manual correction were performed using COOT 51, and structural refinement was carried out with PHENIX refine 52. Data collection and refinement statistics are summarized in Table 2. All figures illustrating the protein structures were generated using PyMOL (Schrödinger).

Phylogenetic analysis and Ancestral state reconstruction

The ATPase domains of representative GHKL superfamily members (MutL, GyrB, MORC, and Hsp90) were aligned using amino acid sequences extracted based on domain boundaries defined by structural correspondence. The multiple sequence alignment was trimmed using trimAl 53 to remove ambiguously aligned regions. Maximum-likelihood phylogenetic inference was performed using IQ- TREE (v3.0.1) 54,55. Ancestral state reconstruction was performed on the fixed maximum-likelihood topology using IQ-TREE under the empirical Bayesian framework implemented with the selected amino acid substitution model. The state of the amino acid residue at the alignment position corresponding to aqMutL Glu32 or aqGyrB Asp52 was examined. Maximum-likelihood phylogenies were visualized and annotated using the Interactive Tree of Life (iTOL) web server 56. Trees were displayed in a rectangular layout without specification of an external outgroup. For visualization purposes, branch lengths were not scaled and were displayed with equal lengths in Fig. 6. The corresponding phylogeny with branch lengths proportional to the inferred evolutionary distances is provided in Supplementary Fig. S5.

eLife Assessment

In this solid work, Fukui et al. re-examined the ATP hydrolysis mechanism in GHKL ATPases, revealing a cooperative role for two conserved acidic residues rather than a single one. This valuable study used a range of biochemical and structural techniques on various mutants from different members of the GHKL ATPase family to test and validate their proposed mechanism. An updated and extended mechanistic model of ATP hydrolysis by this class of enzymes is proposed.

Reviewer #1 (Public review):

Summary:

In this manuscript the applicants study two residues in the GHKL ATPase active site of Aq MutL and GyrB, and argue that the catalytic base function is shared between two conserved acidic residues that are 3 residues apart.

In the manuscript, they generated mutant versions in MutL and GyrB (both ala and the appropriate Asn/Gln version) and performed ATPase analysis. They also generated high resolution crystal structures of the GyrB NTD with AMPPnP for WT and mutants of the two acidic residues. The data show that mutation in either of these residues does not fully kill activity (with the exception of the Alanine mutation of the first of the two, that interferes with ATP (or AMPPnP) binding). When the acidic residues are mutated to Asn/Gln, the catalytic water can still be positioned, and hence these mutants are more active than the Ala mutants. In both cases the double mutation is catalytic dead.

The authors then perform phylogenetic analysis and ancestral gene reconstruction and based on this they argue that HSP90 forms a different class of GHKL ATPases, and lost rather than gained this separate status.

Strengths:

The biochemical analysis seems solid.

Weaknesses:

- A major question that remains, is why the mutations have so much more detrimental effect in MutL (100-fold lower kcat/KM) than they do in GyrB (3-fold lower). Can the authors explain this? Doesn't this argue against the proposed catalytic conservation?

The authors need to discuss this issue explicitly to make it clear that conservation of the mechanism is not complete and that other interpretations are possible.

- The structure figures all have omit maps for just the AMPPnP and the water, whereas the density for the the acidic residues and their mutants are not shown.

This has been addressed.

There are some issues with figure S2B and S5.

Reviewer #2 (Public review):

Summary:

In this manuscript, Fukui et al. re-examined the ATP hydrolysis mechanism in GHKL ATPases, revealing a cooperative role of two conserved acidic residues rather than one. The authors have used a range of biochemical and structural techniques on various mutants from different members of the GHKL ATPase family to test and validate their proposed mechanism.

Through a detailed re-analysis of their previously published structure of the aqMutL NTD (ATPase domain) in complex with AMPPCP, they identified Glu29 and Glu32 as interacting with nucleophilic water for the catalysis. The authors carefully dissected the respective roles of these two acidic residues with a series of site-directed mutations. Mutations at Glu29 impaired ATPase activity without affecting protein secondary structure or ATP binding in the case of the E29Q mutant. Moreover, mutations at Glu32 did not affect secondary structure (except for E32G) but reduce ATPase activity. Activity was abolished when both residues (E29Q/E32Q) are mutated.

The authors extended their study to another GHKL ATPase, aqGyrB. Their findings further supported the cooperative function of the corresponding acidic residues in aqGyrB (Glu48 and Asp51) during ATP hydrolysis. Mutation of these residues partially impaired ATP hydrolysis without affecting protein secondary structure. ATPase activity was completely lost in the double mutant E48Q/D51M. While the E48Q mutant retained the ability to bind ATP, the E48A mutant did not. High-resolution structures of the WT and E48A, E48Q, D51A and D51N mutants of the aqGyrB NTD demonstrated that nucleophilic water positioning depended on these residues. E48 played a dominant role in water positioning and is critical for stabilising ATP lid formation and associated conformational changes, whereas D51 contributed cooperatively to catalysis.

The authors investigated the functional impact of mutating the corresponding residues in the human MutL homologs PMS2 and MLH1. Clinical variants consistently exhibited reduced or abolished ATPase activity, providing a potential molecular basis for Lynch syndrome, through impaired DNA mismatch repair.

Lastly, through evolutionary analysis, the authors inferred that the second acidic residue was likely present in the common ancestor of MutL, GyrB, and MORC proteins, but was lost in the case of Hsp90.

Strengths:

(1) This study contains a detailed structural and biochemical analysis of a biologically important set of GHKL ATPases. The authors identify a second acidic residue that is conserved and contributes to catalysis in a large subset of GHKL ATPases. An updated and extended mechanistic model of ATP hydrolysis by this class of enzymes is proposed, which involves cooperative and partially overlapping roles for the catalytic residue pair. This revised mechanistic model is invaluable for the interpretation of clinical variants of GHKL ATPases such as PMS2 and MLH1.

(2) The work described was performed to an excellent and rigorous technical standard. The structural and biochemical data are sound. The evidence supporting the claims is compelling.

Weaknesses:

(1) The identification in this study of a second acidic residue contributing to catalysis but not absolutely essential for catalysis is a useful finding. However, given that many structures of GHLK ATPases have been determined with different nucleotide analogs bound and that the essential role of the first acidic residue is well established, the importance and scope of the advances described here remain focused within the field of study of GHKL ATPases.

(2) The authors assessed the consequences of variants in the human MutL homologs PMS2 and MLH1, but various other human GHKL ATPases contain clinically relevant variants, some of which have stronger disease associations than the mutations examined in this study. A broader analysis of any effect of disease-linked mutations in GHKL ATPases would have strengthened this study.

(3) The effect of other aqMutL NTD E32 mutants, particularly, the E32K mutant on ATP binding remains unclear, although experimental assessment of nucleotide binding would be challenging due to the high protein concentrations required for the equilibrium dialysis assay.

Author response:

The following is the authors’ response to the previous reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

In this manuscript the applicants study two residues in the GHKL ATPase active site of Aq MutL and GyrB, and argue that the catalytic base function is shared between two conserved acidic residues that are 3 residues apart.

In the manuscript, they generated mutant versions in MutL and GyrB (both ala and the appropriate Asn/Gln version) and performed ATPase analysis. They also generated high resolution crystal structures of the GyrB NTD with AMPPnP for WT and mutants of the two acidic residues. The data show that mutation in either of these residues does not fully kill activity (with the exception of the Alanine mutation of the first of the two, that interferes with ATP (or AMPPnP) binding). When the acidic residues are mutated to Asn/Gln, the catalytic water can still be positioned, and hence these mutants are more active than the Ala mutants. In both cases the double mutation is catalytic dead.

The authors then perform phylogenetic analysis and ancestral gene reconstruction and based on this they argue that HSP90 forms a different class of GHKL ATPases, and lost rather than gained this separate status.

Strengths:

The biochemical analysis seems solid.

Weaknesses:

A major question that remains, is why the mutations have so much more detrimental effect in MutL (100-fold lower kcat/KM) than they do in GyrB (3-fold lower). Can the authors explain this? Doesn't this argue against the proposed catalytic conservation?

The authors need to discuss this issue explicitly to make it clear that conservation of the mechanism is not complete and that other interpretations are possible.

The structure figures all have omit maps for just the AMPPnP and the water, whereas the density for the the acidic residues and their mutants are not shown.

This has been addressed.

There are some issues with figure S2B and S5.

Reviewer #2 (Public review):

Summary:

In this manuscript, Fukui et al. re-examined the ATP hydrolysis mechanism in GHKL ATPases, revealing a cooperative role of two conserved acidic residues rather than one. The authors have used a range of biochemical and structural techniques on various mutants from different members of the GHKL ATPase family to test and validate their proposed mechanism.

Through a detailed re-analysis of their previously published structure of the aqMutL NTD (ATPase domain) in complex with AMPPCP, they identified Glu29 and Glu32 as interacting with nucleophilic water for the catalysis. The authors carefully dissected the respective roles of these two acidic residues with a series of site-directed mutations. Mutations at Glu29 impaired ATPase activity without affecting protein secondary structure or ATP binding in the case of the E29Q mutant. Moreover, mutations at Glu32 did not affect secondary structure (except for E32G) but reduce ATPase activity. Activity was abolished when both residues (E29Q/E32Q) are mutated.

The authors extended their study to another GHKL ATPase, aqGyrB. Their findings further supported the cooperative function of the corresponding acidic residues in aqGyrB (Glu48 and Asp51) during ATP hydrolysis. Mutation of these residues partially impaired ATP hydrolysis without affecting protein secondary structure. ATPase activity was completely lost in the double mutant E48Q/D51M. While the E48Q mutant retained the ability to bind ATP, the E48A mutant did not. High-resolution structures of the WT and E48A, E48Q, D51A and D51N mutants of the aqGyrB NTD demonstrated that nucleophilic water positioning depended on these residues. E48 played a dominant role in water positioning and is critical for stabilising ATP lid formation and associated conformational changes, whereas D51 contributed cooperatively to catalysis.

The authors investigated the functional impact of mutating the corresponding residues in the human MutL homologs PMS2 and MLH1. Clinical variants consistently exhibited reduced or abolished ATPase activity, providing a potential molecular basis for Lynch syndrome, through impaired DNA mismatch repair.

Lastly, through evolutionary analysis, the authors inferred that the second acidic residue was likely present in the common ancestor of MutL, GyrB, and MORC proteins, but was lost in the case of Hsp90.

Strengths:

(1) This study contains a detailed structural and biochemical analysis of a biologically important set of GHKL ATPases. The authors identify a second acidic residue that is conserved and contributes to catalysis in a large subset of GHKL ATPases. An updated and extended mechanistic model of ATP hydrolysis by this class of enzymes is proposed, which involves cooperative and partially overlapping roles for the catalytic residue pair. This revised mechanistic model is invaluable for the interpretation of clinical variants of GHKL ATPases such as PMS2 and MLH1.

(2) The work described was performed to an excellent and rigorous technical standard. The structural and biochemical data are sound. The evidence supporting the claims is compelling.

Weaknesses:

(1) The identification in this study of a second acidic residue contributing to catalysis but not absolutely essential for catalysis is a useful finding. However, given that many structures of GHLK ATPases have been determined with different nucleotide analogs bound and that the essential role of the first acidic residue is well established, the importance and scope of the advances described here remain focused within the field of study of GHKL ATPases.

(2) The authors assessed the consequences of variants in the human MutL homologs PMS2 and MLH1, but various other human GHKL ATPases contain clinically relevant variants, some of which have stronger disease associations than the mutations examined in this study. A broader analysis of any effect of disease-linked mutations in GHKL ATPases would have strengthened this study.

(3) The effect of other aqMutL NTD E32 mutants, particularly, the E32K mutant on ATP binding remains unclear, although experimental assessment of nucleotide binding would be challenging due to the high protein concentrations required for the equilibrium dialysis assay.

We are grateful to the Editors and reviewers for their careful assessment of our revised manuscript and for identifying the remaining points that required clarification. We have addressed each of these comments in the present revision. We believe that these revisions have resolved the remaining concerns and have further improved the clarity and accuracy of the manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Please discuss the large difference in the effect of the mutants on activity explicitly

According to the reviewer’s suggestion, we have added the following discussion to the revised manuscript:

“Although mutation of the two acidic residues impaired the ATPase activity in both aqMutL and aqGyrB, the magnitude of the effects differed substantially, with much greater reduction in the catalytic efficiency in aqMutL than in aqGyrB (Table 1). The molecular basis for this quantitative difference is currently unclear. One possible explanation is that subtle differences in the active-site architecture and surrounding residues alter the relative contribution of each acidic residue to catalysis, allowing aqGyrB to tolerate perturbation of either residue more effectively than aqMutL.” (p.6 line 257-262 in the revised manuscript)

(2) Figure S2B is a completely different view from the other panels, please provide the correct one.

We have revised Supplementary Fig. S2B so that the E48A structure is now shown from a viewpoint as similar as possible to those used in the other panels. We note, however, that the E48A structure cannot appear completely identical to the other panels because the E48A mutant does not bind the ATP analog and therefore does not undergo the nucleotide-binding-associated conformational changes observed in the other structures.

(3) S5 : It is not clear to me what is meant by " The scale bar indicates the number of amino acid substitutions per site." : there is a 'Tree Scale 1" but no other numbers in my version.

We thank the reviewer for pointing out that the scale bar in Supplementary Fig. S5 was insufficiently explained. The value “1” in the tree scale corresponds to a branch length of one amino acid substitution per site. To avoid ambiguity, we have revised the scale-bar label in Supplementary Fig. S5 to explicitly indicate “1 substitution/site” and have clarified its meaning in the figure legend:

“Branch lengths are proportional to the evolutionary distances inferred by IQ-TREE. The scale bar represents an evolutionary distance of one amino acid substitution per site.” (p. 22 line 767-769 in the revised manuscript)

Reviewer #2 (Recommendations for the authors):

(1) P. 9, in the "Data Accessibility Statement", all three PDB codes (23UX, 23UY, and 23UZ) should be listed.

We thank the reviewer for this comment. We carefully rechecked the Data Accessibility Statement and confirmed that all three PDB accession codes (23UX, 23UY, and 23UZ) are included in the statement.

(2) Supplementary Figures S2 and S3. The authors have written "Asn33" and "Asn52", instead of "Glu32" and "Asp51" in both the figure and figure legend of Supplementary Figure S3. They have also written "TND" instead of "NTD" in the figure legend.

“Asn33” and “Asn52” in Supplementary Fig. S3 are not typographical errors. Asn33 in aqMutL and Asn52 in aqGyrB are the residues that directly coordinate the Mg2+ ion and are distinct from the acidic residues discussed in this paper. To avoid confusion, we have added the following sentence to the legend of Supplementary Fig. S3:

“These Mg2+-coordinating asparagine residues are adjacent to, but distinct from, the second acidic residues Glu32 in aqMutL and Asp51 in aqGyrB examined in this study.” (p. 21 line 749-751 in the revised manuscript)

We have also corrected the typographical error “TND” to “NTD” in the figure legend. (p. 21 line 749 in the revised manuscript)

Version 3

Abstract

GHKL ATPases share a unique Bergerat ATP-binding fold and regulate diverse biological processes through ATP-dependent conformational changes. An early step of ATP hydrolysis in this family has been attributed to a single highly conserved glutamate residue proposed to function as the general base. However, mutations of this residue impair both the ATPase activity and ATP binding, complicating interpretation of its catalytic role. Re-examination of the high-resolution crystal structures revealed a second conserved acidic residue positioned within a hydrogen-bonding distance from the nucleophilic water molecule. Using Aquifex aeolicus MutL and GyrB as model enzymes, we combined systematic mutagenesis, ATPase and ATP-binding assays, and X-ray crystallography to dissect the roles of these residues. We show that alignment of the nucleophilic water can be maintained as long as the conserved glutamate retains hydrogen-bonding capability, whereas efficient ATP hydrolysis requires proton-accepting capacity at least one of the two acidic residues. These results indicate that the conserved glutamate primarily governs positioning of the nucleophilic water, while activation of this water for catalysis is achieved through cooperative general base function of the two acidic residues. Extending this framework to human MutL homologs, PMS2 and MLH1, we showed that clinically reported variants of uncertain significance in these DNA mismatch repair proteins substantially reduced the ATPase activity, indicating functional impairment. Together, our findings refine the catalytic mechanism of GHKL ATPases and provide a structural and functional framework for interpreting disease-associated variants in GHKL ATPases. Phylogenetic and ancestral state analysis further indicated that the second acidic residue was likely to be present in the common ancestor of major GHKL ATPase lineages but was later modified in a branch including Hsp90, suggesting evolutionary remodeling of the catalytic mechanism in the branch.

Introduction

The GHKL ATPase family comprises a wide range of proteins involved in diverse cellular processes, including DNA replication, DNA repair, protein homeostasis, chromatin regulation, and signal transduction 1. The GHKL ATPase family includes the following representative members: DNA gyrase B (GyrB), the ATPase subunit of type II topoisomerase that introduces negative supercoils into DNA 2,3; the molecular chaperone Hsp90, which regulates the folding and stability of numerous client proteins 4; histidine kinases of two-component regulatory systems, which function in environmental sensing and signal transduction 5; the DNA mismatch repair protein MutL, which maintains genome stability by correcting replication errors 6,7; and MORC family proteins, which contribute to chromatin organization and epigenetic regulation 8,9. Despite their functional diversity, these proteins share a conserved Bergerat ATP-binding motif and are thought to employ a common mechanism for ATP binding and hydrolysis 1,10-13.

For GHKL ATPases, ATPase activity is essential for protein function. In many cases, ATP binding and hydrolysis drive large-scale conformational changes that are coupled to biological activity. The ATP-bound and ADP-bound states often correspond to distinct oligomeric or domain arrangements, and cycling between these states enables GHKL ATPases to act as molecular switches or motors that coordinate complex cellular processes 14-16. A representative example of such functional coupling is provided by MutL, a key component of the DNA mismatch repair pathway 17-19. MutL consists of an N-terminal ATPase domain and a C-terminal nuclease domain, and ATP binding induces dimerization of the N-terminal domains, leading to rearrangement of the overall architecture and engagement of downstream repair factors 14,20-22. Subsequent ATP hydrolysis, stimulated by interacting proteins and DNA substrates, induces further conformational changes that activate its nuclease function and enable strand incision during mismatch repair 7,23-25. Thus, in MutL, the ATP binding and hydrolysis cycle directly controls large-scale structural transitions that are essential for its biological activity. A detailed understanding of the ATP hydrolysis mechanism is therefore crucial not only for elucidating the molecular basis of fundamental biological phenomena but also for clarifying disease mechanisms associated with dysfunction of GHKL ATPases.

In human, impaired function of MutL proteins compromises the DNA mismatch repair pathway, leading to reduced fidelity of DNA replication. Such defects are a major cause of Lynch syndrome, one of the most common hereditary cancer predisposition syndromes 26-30. Cancers associated with Lynch syndrome exhibit high responsiveness to immune checkpoint inhibitors 31,32, and genetic testing of MutL genes is recommended for diagnosis. However, clinical interpretation is often hampered by the presence of variants of uncertain significance. This limitation largely reflects the insufficient understanding of MutL function at the amino acid level. Variants of uncertain significance have also been identified in the ATPase active site of human MutL homologs, highlighting the importance of elucidating the ATP hydrolysis mechanism of MutL for establishing a molecular basis for genetic diagnosis of Lynch syndrome.

Among GHKL ATPases, GyrB, MutL, and MORC are thought to share particularly similar ATPase mechanisms 11. Within the Bergerat ATP-binding motif, a highly conserved glutamate residue, Glu29 in Aquifex aeolicus MutL (aqMutL), has been proposed to function as a general base that activates the nucleophilic water molecule for ATP hydrolysis 33,34. This proposal is based on observations that substitution of alanine or lysine for this glutamate abolishes the ATPase activity. In addition, a conserved lysine residue, Lys79, in aqMutL, has been suggested to act as a general acid, donating a proton to the reaction intermediate and thereby facilitating cleavage of the γ-phosphate 11. Other conserved basic residues have also been proposed to contribute as Lewis acids that stabilize negatively-charged reaction intermediates 11,34. Collectively, these residues have been proposed to form the core catalytic machinery of GHKL ATPases.

Previously, we determined high-resolution crystal structures of the ATPase domain of aqMutL in complex with ATP analogs, in which a water molecule presumed to be the nucleophile for ATP hydrolysis was clearly observed at a hydrogen-bonding distance to the conserved glutamate Glu29 35. Upon careful re-examination of these structures, we noticed that this same water molecule is also positioned within a hydrogen-bonding distance to a second conserved acidic residue, Glu32 (Figure 1A). Thus, the nucleophilic water is not exclusively coordinated by Glu29, but simultaneously interacts with two acidic residues.

Effects of mutations on the overall structure and the ATPase activity of the aqMutL NTD.

(A) Crystal structure of the aqMutL N-terminal domain (NTD) in complex with the non-hydrolyzable ATP analog AMPPCP. The boxed region indicates the ATP-binding site and is shown enlarged on the right. AMPPCP and the side chains of Glu29 and Glu32 are shown as stick models. Oxygen atoms are shown in red, nitrogen atoms in blue, and phosphorus atoms in orange. The magnesium ion and the water molecule putatively acting as the nucleophile are depicted as green and red spheres, respectively. Hydrogen bonds are indicated by dashed lines, with distances shown. The blue mesh represents an omit Fo–Fc electron density map contoured at 3σ for AMPPNP, the magnesium ion, and the water molecule. (B) Sequence alignment of representative GHKL ATPases. The Bergerat ATP-binding fold is underlined, and the conserved acidic residues corresponding to aqMutL Glu29 and Glu32 are marked with asterisks. Amino acid sequence alignment was performed by CLUSTAL W program 57 and visualized by ESPript3 58. (C) Far-UV CD spectra of the wildtype and mutant forms of the aqMutL NTD were collected and averaged over 10 consecutive scans. (D, E) ATPase activity of the aqMutL NTD was measured under steady-state conditions using a colorimetric assay that detects the release of inorganic phosphate. Apparent rate constants (kapp) were calculated from the measured phosphate concentrations and plotted as a function of ATP concentration. Kinetic parameters were determined by fitting Michaelis–Menten equation to the data. Each data point represents the mean of three independent experiments, with error bars indicating standard deviation. Theoretical curves calculated from the fitted kinetic parameters are overlaid. Data for 0–1 mM substrate concentrations are shown to compare the wildtype from with mutant forms (D). Activities for higher substrate concentrations were measured for some of the mutant forms (E).

Although earlier biochemical studies, including those from our group, demonstrated that replacement of Glu29 with alanine abolishes the ATPase activity, such mutations also severely impair ATP binding 11,33, complicating interpretation of the specific catalytic role of this residue. In this context, the potential contribution of Glu32 to ATP hydrolysis might have been overlooked. Supporting this notion, the residue corresponding to Glu32 is conserved not only in MutL but also in other GHKL ATPases, including GyrB and MORC proteins (Figure 1B). Furthermore, replacements of the corresponding Glu residue with Lys in Escherichia coli MutL and in human MutL homolog 1 (MLH1) have been reported to abolish DNA mismatch repair activity in cell-based or nuclear extract-based assays 36,37.

In this study, we investigated the functional role of Glu32 in the ATP hydrolysis mechanism of MutL and GyrB using a combination of enzymatic and X-ray crystallographic analyses. Our results provide new insights into the catalytic mechanism of GHKL ATPases and refine the current model of ATP hydrolysis for this protein family. By clarifying the molecular basis of ATP hydrolysis in GHKL ATPases, this work also provides a mechanistic framework for interpreting the functional consequences of some clinically observed variants, thereby contributing to the foundation for assessing the impact of variants of uncertain significance.

Results and Discussion

Mutational analysis of Glu29 and Glu32 in the ATPase domain of aqMutL

To examine the contribution of Glu32 to the ATPase activity of the aqMutL NTD and to directly compare its role with that of Glu29, we introduced a series of substitutions at both positions. To exclude the possibility that the observed effects on the ATPase activity arose from global structural perturbations, circular dichroism (CD) spectra were measured for all mutant forms (Figure 1C). Among these, only the E32G mutant form exhibited a marked decrease in ellipticity, suggesting that this substitution disrupts the secondary structure of the protein. In contrast, all other mutant forms displayed CD spectra indistinguishable from that of the wildtype form, indicating that these substitutions do not substantially perturb the overall structure of the aqMutL NTD.

We first analyzed the effects of mutations at Glu29 on the ATPase activity of the aqMutL NTD. The mutation to Gln would selectively impair the proton association/dissociation capability of the Glu side chain while preserving its size and hydrogen-bonding potential. Moreover, because mutations to Val, Lys, and Gly at this position have been reported as variants of uncertain significance in human MutL, these mutations were also analyzed. Consistent with previous reports, all Glu29 mutant forms exhibited severe reduction in the ATPase activity (Figure 1D and E, and Table 1). In particular, the E29A, E29V, E29K, and E29G mutant forms showed activities below the detection limit, indicating that Glu29 plays a central role in ATP hydrolysis. However, as described above, previous studies also demonstrated that mutations at Glu29 impair ATP binding 11,33,34, indicating that the loss of the ATPase activity of these mutant forms cannot be unambiguously attributed to disruption of a catalytic base function.

To further dissect the role of Glu29, we focused on the E29Q mutant form. Using equilibrium dialysis, we confirmed that the E29Q mutant form binds ATP with an affinity comparable to that of the wildtype protein (Supplementary Figure S1A). Because this substitution does not impair ATP binding, it should enable a more direct assessment of the proposed role of Glu29 as a general base in ATP hydrolysis. The E29Q mutant form retained measurable ATPase activity, although its kcat value was approximately tenfold smaller than that of the wildtype enzyme (Figure 1D and E, and Table 1). The persistence of substantial catalytic activity in the absence of the carboxylate group at the position 29 suggests that a residue other than Glu29 can participate in proton abstraction from the nucleophilic water during ATP hydrolysis.

We next examined the effects of mutations at Glu32. In contrast to the E29A mutant form of the aqMutL NTD, the E32A mutant form exhibited ATP binding ability comparable to that of the wildtype form (Supplementary Figure S1A), indicating that Glu32 does not contribute to ATP binding. Mutations of Glu32 resulted in reduced ATPase activity relative to the wildtype form, supporting a role for this residue in catalysis. However, with the exception of the E32K and the structurally perturbed E32G mutant forms, all Glu32 mutant forms retained detectable ATPase activity (Figure 1D and E, and Table 1), indicating that loss of Glu32 alone is insufficient to abolish ATP hydrolysis. Simultaneous substitutions of glutamine for the two residues in the E29Q/E32Q double mutant form led to a complete loss of the ATPase activity. Taken together, these results indicate that Glu29 and Glu32 do not exert equivalent functions but act cooperatively. The presence of at least one carboxylate group at either position, 29 or 32, appears to be required for ATP hydrolysis.

Conserved roles of two acidic residues of aqGyrB in ATP hydrolysis

To examine whether the cooperative roles of Glu29 and Glu32 observed in aqMutL are conserved among other GHKL ATPases, we next analyzed the NTD of A. aeolicus GyrB (aqGyrB). A glutamate residue corresponding to aqMutL Glu29 is also conserved in bacterial GyrB and has been proposed to function as a general base in ATP hydrolysis 38. Although crystal structure of the E. coli GyrB NTD has been reported 39, the position of the putative nucleophilic water has not been clearly defined. Therefore, we selected the GyrB NTD from the hyperthermophilic bacterium A. aeolicus as a model system, anticipating that it would allow high-resolution structural analysis and clearer visualization of a catalytic water molecule.

First, we determined the crystal structure of this protein in complex with AMPPNP at 1.65 Å resolution (Table 2 and Figure 2A). The overall structure was highly similar to that of the E. coli GyrB NTD, with a Cα root mean square deviation of 1.42 Å. As observed for the E. coli GyrB NTD, AMPPNP was accommodated by the Bergerat ATP-binding fold, and AMPPNP binding seemed to induce conformational changes that promoted dimer formation through the newly formed interface. Glu48 and Asp51 of aqGyrB correspond to Glu29 and Glu32 of aqMutL, respectively. In the crystal structure, these residues were found to tightly coordinate a water molecule that is likely to act as the nucleophilic water during ATP hydrolysis (Figure 2A). Based on these structural observations, we introduced Ala or Gln/Asn substitutions for Glu48 and Asp51. The CD spectra of the mutant forms were nearly identical to that of the wildtype form (Figure 2B), suggesting that these substitutions did not disrupt the overall structure of the aqGyrB NTD.

Crystal structure of the aqGyrB NTD and effects of mutations on the overall structure and the ATPase activity of the aqGyrB NTD.

(A) Structure of the AMPPNP-bound dimer of the wildtype aqGyrB NTD. The boxed region highlights the ATP-binding site and is enlarged in the left panel. The same region is further enlarged and rotated 60° to provide an alternative view of the active site geometry. AMPPNP and the side chains of Glu48 and Asp51 are shown as stick models. Oxygen atoms are shown in red, nitrogen atoms in blue, and phosphorus atoms in orange. The magnesium ion and the water molecule presumed to act as the nucleophile are depicted as green and red spheres, respectively. Hydrogen bonds are indicated by dashed lines, with distances shown. The blue mesh represents an omit Fo–Fc electron density map contoured at 3σ for AMPPNP, the magnesium ion, and the water molecule. (B) Far-UV CD spectra of the wildtype and mutant forms of the aqGyrB NTD were recorded and averaged over 10 accumulations. (C) ATPase activity of the aqGyrB NTD was measured under steady-state conditions as described in Figure 1. The kapp values were plotted as a function of ATP concentration. Michaelis–Menten equation was fitted to the data. Data points represent the mean ± standard deviations from three independent experiments.

Equilibrium dialysis showed that the E48A mutant form failed to bind ATP, whereas the E48Q mutant form displayed ATP-binding affinity comparable to that of the wildtype protein (Supplementary Figure S1B). The E48Q mutation therefore allowed us to evaluate the catalytic contribution of Glu48 independently of ATP binding. ATPase activity assays showed that the E48A mutant form completely lacked the activity, while the E48Q mutant form retained approximately one third of the wildtype activity (Figure 2C and Table 1). The persistence of substantial catalytic activity in the E48Q mutant form suggests that another residue partially compensates for the loss of the Glu48 carboxylate in the proton abstraction.

The D51A mutant form of the aqGyrB NTD retained ATP binding ability comparable to that of the wildtype form, indicating that Asp51 is not required for nucleotide binding (Supplementary Figure S1B). Mutations of Asp51 also impaired ATP hydrolysis, with the D51A and D51N mutant forms retaining ∼16% and ∼50% of the wildtype activity, respectively (Figure 2C and Table 1). Notably, the E48Q/D51N double mutant form exhibited no detectable ATPase activity, resembling the effect of the corresponding double mutation in the aqMutL NTD. Taken together, these results indicate that both Glu48 and Asp51 contribute to the catalytic mechanism of ATP hydrolysis in aqGyrB and that at least one of these residues must retain a carboxylate group with the proton-accepting capability to sustain the enzymatic activity.

Crystal structures of aqGyrB mutant forms and the role of Glu48 in ATP binding

In order to relate the ATPase activity profiles of the mutant forms to their structural features, we determined X-ray crystal structures of selected mutant forms using the aqGyrB NTD as a model system, which allows relatively high-resolution structural analyses. The AMPPNP-bound structures of the E48A, E48Q, D51A, and D51N mutant forms were determined at resolutions of 2.66, 1.70, 1.63, and 2.05 Å, respectively (Table 2). To show that the introduced substitutions were unambiguously supported by the crystallographic data, electron density maps around residues 48 and 51 are shown in Supplementary Figure S2.

Structural superposition of the aqGyrB NTD and aqMutL NTD revealed that the catalytic Mg² ion occupies essentially the same position in the two ATPase active sites (Supplementary Figure S3), indicating that the metal-binding geometry is highly conserved, where the Mg2+ ion is coordinated by the side chain of Asn52, AMPPNP, and surrounding water molecules. Neither Glu48 nor Asp51 directly coordinated the Mg2+ ion.

Since the conserved general base Glu (i.e. Glu48 of aqGyrB) does not directly interact with ATP, it has been unclear why mutations at the Glu residue abolish the ATP-binding ability of GHKL ATPases. To clarify this, we compared the crystal structures of the wildtype and mutant forms of the aqGyrB NTD (Figure 3). In GHKL ATPases, ATP binding is tightly coupled to conformational rearrangements, including formation of an ATP lid (Figure 3A), which acts to secure the bound nucleotide by enclosing it within the active site 3,21. Structural analysis of the wildtype aqGyrB NTD revealed that the side chain of Glu48 forms a hydrogen bond with Gln340, thereby stabilizing a loop containing Gln340 in a conformation capable of ATP binding (Figure 3B). In this arrangement, Gln340 in turn forms a hydrogen bond with His122, a residue located within the ATP lid, while Lys342 in the same loop is positioned to interact with the γ-phosphate of ATP. These coordinated interactions appear to promote the ATP-induced conformational transition required for stable nucleotide binding.

Structural basis for the loss and preservation of ATP binding in aqGyrB NTD mutant forms.

The wildtype and all mutant forms were crystallized in the presence of AMPPNP; however, the E48A mutant form did not bind AMPPNP, whereas the E48Q, D51A, and D51N mutant forms were observed in complex with AMPPNP. (A) Overall structures of the wildtype and E48A mutant aqGyrB NTD monomers. The ATPase active site is indicated by a boxed region. The ATP-lid (residues 105–124) in the wildtype form and the corresponding region in the E48A mutant form are shown in magenta and olive, respectively. (B–F) Enlarged views of the ATPase active site in the wildtype (B), E48A (C), E48Q (D), D51A (E), and D51N (F) forms. Side chains of the residues implicated in the ATP-induced conformational changes (e.g. His122, Gln340, and Lys342), as well as residues 48 and 51, are shown as stick models. AMPPNP is depicted as sticks, and the Mg² ion is shown as a green sphere. Hydrogen bonds and ionic interactions are indicated by dashed lines.

In contrast, structural analysis of the E48A mutant form demonstrated that disruption of the interaction with Gln340 abolishes this network of hydrogen bonds, preventing the associated conformational rearrangements and thereby impairing ATP binding (Figure 3C). These findings indicate that the interaction network emanating from Glu48 is indispensable for the conformational changes associated with the ATP lid formation. Interestingly, in the E48Q mutant form, the hydrogen bond with Gln340 is preserved, allowing the conformational transition to occur and enabling ATP binding (Figure 3D). Consistently, in the D51A and D51N mutant forms, the interaction mediated by Glu48 remains intact, and the structural rearrangements required for ATP binding are maintained (Figure 3E and F).

These structural observations extend previous models of GyrB ATPase function by linking the conserved catalytic glutamate to ATP lid stabilization and nucleotide-dependent conformational rearrangement. The E48A aqGyrB NTD structure, in particular, provides direct structural evidence that disruption of this Glu-centered interaction network is sufficient to prevent ATP-induced active-site assembly. This hydrogen-bonding network emanating from the general base Glu is conserved in GyrBs from other organisms 39,40 (Supplementary Figure S4), while it is not observed in other GHKL ATPases, including MutL 22,34,35,41 and MORC family proteins 9,42,43. Therefore, the loss of ATP-binding ability observed upon the mutation of Glu29 to Ala in MutL 11 is likely to be explained by a different mechanism. This difference highlights mechanistic diversity within the GHKL ATPase family, despite conservation of the two catalytic acidic residues.

Nucleophilic water molecule in the active site of mutant forms of the aqGyrB NTD

Having established the structural basis for the altered ATP-binding properties of the mutant forms, we next analyzed how these substitutions influence the positioning of the nucleophilic water molecule. Structural comparisons of the catalytic sites of the four mutant forms are shown in Figure 4. In the E48A mutant form (Figure 4A), the nucleophilic water molecule was not observed in the catalytic site. In contrast, in the E48Q mutant form, the nucleophilic water molecule was positioned at essentially the same location as in the wildtype form, coordinated by Gln48 and Asp51, and bound AMPPNP was observed (Figure 4B). This structural arrangement is consistent with the biochemical data described above, in which the E48Q mutant form retained substantial ATP-binding and ATPase activities.

Structures surrounding the conserved acidic residues: (A) E48A, (B) E48Q, (C) D51A, and (D) D51N mutant forms of the aqGyrB NTD.

AMPPNP and the side chains at the positions 48 and 51 are shown as stick models. Oxygen atoms are shown in red, nitrogen atoms in blue, and phosphorus atoms in orange. The magnesium ion is shown as a green sphere, and the water molecule proposed to act as the nucleophilic water is shown as a red sphere. Blue mesh represents Fo–Fc omit electron density maps contoured at 3σ, calculated after omission of AMPPNP, the magnesium ion, and the nucleophilic water molecule. Hydrogen bonds are indicated by dashed lines, with the corresponding distances shown.

Unlike the E48A mutant form, the D51A mutant form bound AMPPNP (Figure 4C), and the nucleophilic water molecule was coordinated only by Glu48. This asymmetric coordination suggests that Glu48 plays a dominant role in positioning the nucleophilic water molecule. In the D51N mutant form (Figure 4D), which lacks the proton-transfer capability of Asp51 while retaining its hydrogen-bonding capacity, the nucleophilic water was coordinated by both Glu48 and Asn51. These structural features are consistent with the relative ATPase activities observed for the mutant forms of these two acidic residues (Figure 2C).

Taken together, these results indicate that Glu48 and Asp51 make distinct yet cooperative contributions to ATP hydrolysis. Glu48 appears to play a dominant role in aligning the nucleophilic water molecule and activating this water molecule through proton abstraction as a general base. In contrast, Asp51 seems to contribute less to the precise positioning of the nucleophilic water but participates cooperatively with Glu48 in the proton abstraction, thereby supporting efficient ATP hydrolysis. Although mutation of the two acidic residues impaired ATPase activity in both aqMutL and aqGyrB, the magnitude of the effects differed substantially, with a much greater reduction in catalytic efficiency in aqMutL than in aqGyrB. The molecular basis for this quantitative difference is currently unclear. One possible explanation is that subtle differences in the active-site architecture and surrounding residues alter the relative contribution of each acidic residue to catalysis, allowing aqGyrB to tolerate perturbation of either residue more effectively than aqMutL.

Notably, an acidic residue corresponding to Asp51 of aqGyrB is conserved among several GHKL ATPases, including MutL, GyrB, and MORC family proteins (Figure 1B), suggesting that the two-acidic-residue mechanism proposed here may be shared among these subgroups. The second acidic residue is not conserved in other members of the GHKL family, such as Hsp90 and mitochondrial histidine kinases (Figure 1B), where a single acidic residue is thought to fully function as the general base responsible for water activation. A similar divergence appears to apply to the previously-reported general acid Lys residue 11, further supporting the notion that GHKL ATPases can be divided into at least two mechanistically distinct subclasses. This mechanistic bifurcation provides insights into the molecular evolution of the GHKL ATPase family.

Functional and clinical implications of the second acidic residue in human MutL homologs

Our analyses of the aqMutL NTD and aqGyrB NTD indicate that, in addition to the Glu residue traditionally considered to function as the general base, an additional conserved acidic residue (Glu or Asp) also participates in the general base catalysis. In human MutL homologs, this residue corresponds to Glu44 in Postmeiotic Segregation Increased 2 (PMS2) and Glu37 in MLH1. Notably, variants of uncertain clinical significance, including E44V (ClinVar variation ID: 232732), E44G (230139), and E44Q (186069) in PMS2, as well as E37G (230139) and E37K (89640) in MLH1, have been reported at these positions.

To examine whether these residues similarly contribute to the ATPase activity in human MutL homologs, we introduced the corresponding mutations into the NTDs of human PMS2 and MLH1 and measured their ATPase activities. The recombinant E44G human PMS2 NTD and E37G human MLH1 NTD could not be expressed successfully, which is consistent with our finding that the corresponding mutation in the aqMutL NTD disrupted the structural integrity of the protein (Figure 1C). Both the E44V and E44Q mutant forms of the PMS2 NTD exhibited a marked reduction in the ATPase activity, retaining approximately one-sixth of the wildtype activity (Figure 5A and Table 1). In contrast, the E37K mutation in the MLH1 NTD completely abolished the ATPase activity under our assay conditions (Figure 5B and Table 1) unlike the corresponding glutamine substitutions, which retained substantial residual ATPase activity in the aqMutL, aqGyrB, and PMS2 NTDs. A similar complete loss of ATPase activity was also observed for the E32K mutant form of the aqMutL NTD. These observations suggest that the severe defect caused by the lysine substitution cannot be attributed simply to loss of the catalytic carboxylate. Instead, introduction of a positively charged side chain (charge reversal) is likely to perturb the local electrostatic environment. Structural characterization of the MLH1 E37K and aqMutL E32K mutant forms will be required to clarify the molecular basis of this severe functional defect.

ATPase activities of human MutL homolog NTDs.

(A) ATPase activity of mutant forms of the human PMS2 NTD. (B) ATPase activity of a mutant form of the human MLH1 NTD. ATPase activities were measured using the same procedures as described for Figure 1 and Figure 2. Apparent rate constants (kapp) were plotted as a function of substrate concentration. Data points represent the mean values from three independent measurements, with error bars indicating standard deviations. Solid lines represent the theoretical Michaelis-Menten curves.

These results are in good agreement with our observations for the aqMutL NTD and aqGyrB NTD, further supporting the notion that this second acidic residue plays a conserved and functionally significant role in ATP hydrolysis among some members of the GHKL ATPase family. Previous studies demonstrated that even a several-fold reduction in the MutL ATPase activity could severely compromise mismatch repair efficiency in vivo 11. In this context, the substantial reduction in the ATPase activity caused by the E44V and E44Q mutations in the PMS2 NTD is likely to impair mismatch repair in human cells. In individuals carrying these germline variants, subsequent loss or inactivation of the remaining wildtype allele would leave only the ATPase-defective MutL protein, thereby compromising mismatch repair and promoting tumorigenesis. Nevertheless, whether these variants indeed compromise mismatch repair activity in vivo remains to be directly tested. Functional validation using cell-based or nuclear extract-based mismatch repair assays, such as the 6-thioguanine sensitivity assay 44 or CIMRA assay 36, will be required to establish a definitive link between the reduced ATPase activity and mismatch repair deficiency in human cells.

Although we focused here on pathogenic variants in the MutL homologs MLH1 and PMS2, extending similar structural and biochemical analyses to disease-associated variants in other human GHKL ATPases will be important for evaluating the generality and clinical relevance of the conserved catalytic mechanism proposed in this study.

Phylogenetic analysis of GHKL ATPases

The newly-identified functionally-important acidic residue is not uniformly conserved across the GHKL superfamily. We therefore used this position as an evolutionary marker to investigate patterns of molecular divergence within the GHKL ATPase domain. In histidine kinases, even the canonical catalytic glutamate is typically replaced, and the ATPase-associated catalytic machinery is not maintained 1,45,46. Therefore, phylogenetic reconstruction was restricted to other GHKL ATPase family members MutL, GyrB, MORC, and Hsp90.

Phylogenetic analysis based on the trimmed ATPase-domain alignment recovered these major GHKL ATPase lineages (MutL, GyrB, MORC, and Hsp90) as a well-supported cluster, indicating relatively close evolutionary relationships among these groups, which is consistent with a previous study 1 (Supplementary Figure S5). Ancestral state reconstruction performed on the maximum-likelihood phylogeny strongly supports the presence of an acidic residue at the position corresponding to aqMutL Glu32 or aqGyrB Asp51 in the internal node uniting the MutL, GyrB, MORC, and Hsp90 clades with a posterior probability of 0.983 (Figure 6). In contrast, the ancestral node of the Hsp90 clade is strongly supported to encode serine at this position with a posterior probability of 1.000. Because no external outgroup was specified, the global polarity of amino acid residue changes across the entire phylogeny cannot be formally determined. Nevertheless, the widespread conservation of the second acidic residue across the MutL, GyrB, and MORC lineages, together with the high posterior probability at the most recent common ancestor of these groups and Hsp90, indicates that this residue was very likely to be present in their ancestor. Its absence within the Hsp90 lineage is therefore most parsimoniously interpreted as lineage-specific loss. In summary, these results suggest that the Hsp90 lineage underwent remodeling of its ATPase catalytic mechanism during its evolutionary divergence from other GHKL ATPase family members.

Maximum-likelihood phylogeny of the ATPase domain of representative GHKL family members.

The state of the amino acid residue at the position corresponding to aqMutL Glu32 or aqGyrB Asp51 is indicated by colored bars (red, acidic residue; blue, non-acidic residue). The phylogenetic tree was visualized and annotated using the Interactive Tree of Life (iTOL) web server 56. The tree is displayed in a rectangular layout for clarity and should be interpreted as unrooted. Ancestral state reconstruction was performed on the fixed topology, and posterior probabilities are shown for selected internal nodes. The internal node uniting the MutL/GyrB/MORC/Hsp90 clades strongly supports an acidic residue with the posterior probability (PP) of 0.983, whereas the Hsp90 stem ancestor is strongly supported to encode serine at this position with the posterior probability of 1.000. Species abbreviations used in the phylogenetic tree are as follows: Ce, Caenorhabditis elegans; Sc, Saccharomyces cerevisiae; Dr, Danio rerio; Hs, Homo sapiens; Aa, Aquifex aeolicus; Hp, Helicobacter pylori; Tt, Thermus thermophilus; Ec, Escherichia coli; Mt, Mycobacterium tuberculosis; Bs, Bacillus subtilis; Sm, Streptococcus mutans; Dm, Drosophila melanogaster; Clf, Canis lupus familiaris; Mm, Mus musculus; Hc, Hemicordylus capensis; Pv, Patella vulgata; Tf, Tachysurus fulvidraco; Lk, Lepidochelys kempii; Cc, Castor canadensis; Dn, Dasypus novemcinctus.

Conclusions

In this study, we re-examined the catalytic mechanism of GHKL ATPases using aqMutL and aqGyrB as model systems with mutational, biochemical, and crystallographic studies. Our results reveal that ATP hydrolysis in these enzymes is not governed by a single general base residue, as has been traditionally assumed, but instead relies on the coordinated action of two conserved acidic residues. Structural analyses indicate that Glu48 of aqGyrB plays a dominant role in aligning the nucleophilic water molecule in an optimal geometry for attack on the γ-phosphate of ATP. However, biochemical and structural evidence from single and double mutant forms indicates that both residues contribute to proton abstraction from the nucleophilic water. Thus, general base catalysis in GyrB and MutL appears to be achieved by redundant proton-accepting residues.

Extending this finding to human MutL homologs, we showed that mutations at the corresponding residues in PMS2 and MLH1—reported as variants of uncertain clinical significance—lead to substantial or complete loss of the ATPase activity. Given that even modest reductions in the MutL ATPase activity are known to severely compromise mismatch repair efficiency in vivo, our findings suggest that these variants might be functionally deleterious. Thus, the mechanistic insights obtained in this study provide a structural and biochemical basis for evaluating the pathogenic potential of clinically ambiguous variants in human MutL homologs.

Materials and Methods

Construction of expression plasmids

Expression plasmids encoding mutant forms of the aqMutL NTD (residues 1–315), the aqGyrB NTD (residues 1–397), the human PMS2 NTD (residues 1–365), and the human MLH1 NTD (residues 1–337) were generated by introducing point mutations into pET-11a/aqMutL NTD 47, pET-11a/aqGyrB NTD 11, pCold ProS2/human PMS2 NTD 35, and pET28-MLH1/human MLH1 NTD11 plasmids using the PrimeSTAR site-directed mutagenesis protocol (Takara). Primer sequences used for mutagenesis are listed in Supplementary Table S1. DNA sequencing confirmed that no unintended mutations were introduced.

Expression and purification of proteins

The wildtype aqMutL NTD, aqGyrB NTD, ProS2-tagged human PMS2 NTD, and His-tagged human MLH1 NTD were overexpressed and purified as described previously 11,35,47. The mutant forms were overexpressed and purified using the same procedures as for the respective wildtype proteins.

CD spectrometry

CD measurements were performed with a spectropolarimeter, model J-720W (Jasco). Measurements of CD spectra were carried out in a solution comprised of 20 mM Tris-HCl (pH 8.0) (eLANT) and 10 μM protein using a 0.1 cm cell at 25°C. The residue molar ellipticity [Θ] was defined as 100 Θobs/(lc), where Θobs is the observed ellipticity, l is the length of the light path in centimeters, and c is the residue molar concentration of the protein.

ATPase assay

ATPase activities of the aqMutL NTD, aqGyrB NTD, ProS2-tagged human PMS2 NTD, and His-tagged human MLH1 NTD were measured as described previously 35. Reactions were carried out using 10 μM protein in a 20 μl reaction mixture containing 50 mM HEPES–KOH (pH 7.5) (eLANT), 100 mM NaCl, 5 mM MgCl2, and various concentrations of ATP (Roche). For the aqMutL NTD and aqGyrB NTD, reactions were performed at 70°C for 60 min. For the NTDs of the human MutL homologs, reactions were performed at 37°C for 120 min. All experiments were conducted in triplicate. Kinetic parameters were determined by fitting the data to the standard Michaelis–Menten equation using IgorPro.

Equilibrium dialysis

Equilibrium dialysis experiments were performed as described previously 11. Seventy-five microliters of buffer containing 50 mM Tris–HCl (pH 8.0), 100 mM NaCl, 5 mM MgCl2, and 50 μM AMPPNP (Roche) were loaded into the buffer chamber of a DispoEquilibrium Dialyzer equipped with a 10 kDa molecular weight cut-off membrane (Harvard Bioscience, Inc.). An equal volume of the aqMutL NTD (0–320 μM) or aqGyrB NTD (0–400 μM) in the same buffer without AMPPNP was placed in the opposing sample chamber. The device was gently agitated at room temperature for 24 h to allow equilibration. The absorbance of the solution in the buffer chamber was measured at 260 nm. The concentration of unbound AMPPNP was calculated using a molar extinction coefficient (ε = 15,400 M-1 cm-1).

Crystallization and structure determination

Protein crystallization was performed by the sitting-drop vapor diffusion method at 20°C. One microliter of the aqGyrB NTDs (14 mg/ml) containing 0.6 mM AMPPNP (Roche) and 1 mM MgCl was mixed with an equal volume of the reservoir solution and equilibrated against 500 μl of reservoir solution. For the wildtype and D51A mutant forms of the aqGyrB NTD, crystals were obtained using a reservoir solution containing 0.2 M ammonium acetate, 0.1 M HEPES (pH 7.5), and 45% (v/v) 2-methyl-2,4-pentanediol. The E48A mutant form was crystallized in 0.05 M cadmium sulfate hydrate, 0.1 M HEPES (pH 7.5), and 20% (w/v) polyethylene glycol 3,350. The E48Q mutant form was crystallized under identical conditions consisting of 30% (w/v) polyethylene glycol 400, 100 mM HEPES–NaOH (pH 7.5), and 200 mM sodium chloride. The D51N mutant form was crystallized in 0.1 M sodium malonate (pH 7.0) and 12% (w/v) polyethylene glycol 3,350. Crystals of the wildtype and E48A mutant forms were soaked in the respective reservoir solutions supplemented with 30% (v/v) glycerol before being flash-cooled in liquid nitrogen. Crystal of the D51N mutant form was transferred to the reservoir solution containing 30% (w/v) polyethylene glycol 3,350 prior to flash-cooling in liquid nitrogen. Crystal of the D51A mutant form was directly flash-cooled in liquid nitrogen without additional cryoprotection.

X-ray diffraction data were collected at beamline BL45XU in SPring-8 (Hyogo, Japan) at a wavelength of 1.000 Å and a temperature of −173°C. Data were automatically collected and processed using the ZOO system 48, in which data processing was assisted by KAMO 49. The linearity of the Wilson plot was confirmed over the resolution range employed. Phases were determined by molecular replacement using Phaser with a search model predicted by AlphaFold2 50. Model building and manual correction were performed using COOT 51, and structural refinement was carried out with PHENIX refine 52. Data collection and refinement statistics are summarized in Table 2. All figures illustrating the protein structures were generated using PyMOL (Schrödinger).

Phylogenetic analysis and Ancestral state reconstruction

The ATPase domains of representative GHKL superfamily members (MutL, GyrB, MORC, and Hsp90) were aligned using amino acid sequences extracted based on domain boundaries defined by structural correspondence. The multiple sequence alignment was trimmed using trimAl 53 to remove ambiguously aligned regions. Maximum-likelihood phylogenetic inference was performed using IQ-TREE (v3.0.1) 54,55. Ancestral state reconstruction was performed on the fixed maximum-likelihood topology using IQ-TREE under the empirical Bayesian framework implemented with the selected amino acid substitution model. The state of the amino acid residue at the alignment position corresponding to aqMutL Glu32 or aqGyrB Asp52 was examined. Maximum-likelihood phylogenies were visualized and annotated using the Interactive Tree of Life (iTOL) web server 56. Trees were displayed in a rectangular layout without specification of an external outgroup. For visualization purposes, branch lengths were not scaled and were displayed with equal lengths in Figure 6. The corresponding phylogeny with branch lengths proportional to the inferred evolutionary distances is provided in Supplementary Figure S5.